---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-10-01

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
# devtools::install_git("https://github.com/Gulf-IEA/IEAnalyzeR")
# devtools::install_github("https://github.com/Gulf-IEA/IEAnalyzeR/tree/feat_NoLineMissData", force = T)

library(IEAnalyzeR)
library(here)
library(ggplot2)
library(dplyr)
library(tidyr)

```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "mmtd_bycatch"

csv_filename_otter<-here(paste0("data/formatted/formatted_csvs/", paste0(root_name, "_otter"), "_formatted.csv")) 
object_filename_otter<-here(paste0("data/formatted/final_objects/", paste0(root_name, "_otter"), "_object.rds"))
plot_filename_otter<-here(paste0("figures/plots/", paste0(root_name, "_otter"), "_plot.png"))

csv_filename_try<-here(paste0("data/formatted/formatted_csvs/", paste0(root_name, "_try"), "_formatted.csv")) 
object_filename_try<-here(paste0("data/formatted/final_objects/", paste0(root_name, "_try"), "_object.rds"))
plot_filename_try<-here(paste0("figures/plots/", paste0(root_name, "_try"), "_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

Breaking Down what tables are:
Table 3: Bycatch & Mortality for GoA otter trawl (updated method) 2016-2020
Table 4: Bycatch & Mortality for GoA try nets 2016-2020

```{r}
tab3_otter<-data.frame(year=rep(2016:2020, 2),
                       type=rep(c("Bycatch", "Mortality"), each=5),
                       kemps=c(113,257,228,411,393,38,85,75,136,127),
                       kemps_low95=c(19,47,38,62,75,6,15,12,19,23),
                       kemps_high95=c(646,1162,1029,2230,1853,206,395,359,809,610),
                       leatherback=c(12,23,22,20,20,4,8,7,7,7),
                       leatherback_low95=c(2,5,5,3,3,1,1,2,1,1),
                       leatherback_high95=c(54,97,79,74,79,18,33,28,25,28),
                       loggerhead=c(22,46,40,35,35,7,15,13,12,12),
                       loggerhead_low95=c(5,11,10,8,8,2,4,3,2,3),
                       loggerhead_high95=c(82,156,146,135,124,29,55,53,47,41),
                       green=c(150,156,144,133,141,50,51,48,45,47),
                       green_low95=c(57,59,52,54,54,18,19,17,16,16),
                       green_high95=c(355,352,360,301,337,127,119,127,113,117),
                       unknown=c(78,83,78,76,76,26,28,26,25,25),
                       unknown_low95=c(21,25,23,25,24,7,8,8,8,8),
                       unknown_high95=c(197,211,198,195,189,70,72,69,64,68))

tab3_otter_noconf<-select(tab3_otter, year, type, kemps, leatherback, loggerhead, green, unknown)

tab3_otter_no_wide<-pivot_wider(tab3_otter_noconf, names_from = type, values_from = c(kemps, leatherback, loggerhead, green, unknown))
```


```{r}
tab4_try<-data.frame(year=rep(2016:2020, 2),
                       type=rep(c("Bycatch", "Mortality"), each=5),
                       kemps=c(200,219,198,181,194,0,0,0,0,0),
                       kemps_low95=c(54,59,48,47,59,0,0,0,0,0),
                       kemps_high95=c(535,588,516,481,454,2,2,1,2,1),
                       loggerhead=c(291,298,275,284,326,0,0,0,0,0),
                       loggerhead_low95=c(100,102,98,95,104,0,0,0,0,0),
                       loggerhead_high95=c(762,766,733,741,968,3,2,2,2,3),
                       green=c(42,43,39,39,43,0,0,0,0,0),
                       green_low95=c(13,13,11,12,15,0,0,0,0,0),
                       green_high95=c(107,121,112,107,114,0,0,0,0,0),
                       unknown=c(63,64,58,57,64,0,0,0,0,0),
                       unknown_low95=c(22,24,20,22,23,0,0,0,0,0),
                       unknown_high95=c(153,169,143,151,185,1,1,1,1,1))

tab4_try_noconf<-select(tab4_try, year, type, kemps, loggerhead, green, unknown)

tab4_try_no_wide<-pivot_wider(tab4_try_noconf, names_from = type, values_from = c(kemps, loggerhead, green, unknown))
```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).
Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.

Otter Trawl Data
```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = c(rep(c("kemps", "leatherback", "loggerhead", "green", "unknown"), each=2))
unit_names = c(rep(c("Bycatch", "Mortality"), 5))
extent_names = c(rep("Gulf of America", 10))

formatted_data_otter = IEAnalyzeR::convert_cleaned_data(tab3_otter_no_wide, indicator_names, unit_names, extent_names)

```

```{r}
#Otter Trawl CI
lowerCI<-select(tab3_otter, year, type, contains("low"))
lowerCI_wd<-pivot_wider(lowerCI, id_cols=year, names_from = type, values_from = !c(year, type))

upperCI<-select(tab3_otter, year, type, contains("high"))
upperCI_wd<-pivot_wider(upperCI, id_cols=year, names_from = type, values_from = !c(year, type))


ci_df_otter<-convert_ci_data(lowerCI_wd, upperCI_wd, indicator_names, unit_names, extent_names)

```

Try Data
```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = c(rep(c("kemps", "loggerhead", "green", "unknown"), each=2))
unit_names = c(rep(c("Bycatch", "Mortality"), 4))
extent_names = c(rep("Gulf of America", 8))

formatted_data_try = IEAnalyzeR::convert_cleaned_data(tab4_try_no_wide, indicator_names, unit_names, extent_names)

```

```{r}
#Try CI
lowerCI<-select(tab4_try, year, type, contains("low"))
lowerCI_wd<-pivot_wider(lowerCI, id_cols=year, names_from = type, values_from = !c(year, type))

upperCI<-select(tab4_try, year, type, contains("high"))
upperCI_wd<-pivot_wider(upperCI, id_cols=year, names_from = type, values_from = !c(year, type))


ci_df_try<-convert_ci_data(lowerCI_wd, upperCI_wd, indicator_names, unit_names, extent_names)

```


# 3. Save Formatted data as csv

This will save your data to the appropriate folder.
```{r}
write.csv(formatted_data_otter, file = csv_filename_otter, row.names = F)

write.csv(formatted_data_try, file = csv_filename_try, row.names = F)
```


# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.
```{r}

data_obj_otter<-data_prep(formatted_data_otter, trends = F, subind = "unit", ci_df = ci_df_otter)

data_obj_try<-data_prep(csv_filename_try, trends = F, subind = "unit", ci_df = ci_df_try)

```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.
```{r}
saveRDS(data_obj_otter, file = object_filename_otter)

saveRDS(data_obj_try, file = object_filename_try)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
plot_otter<-IEAnalyzeR::plot_fn_obj(df_obj = data_obj_otter, trend = F, facet_grid = c("indicator", "unit"),
            manual_ylab = "Species", manual_title = NA)

plot_try<-IEAnalyzeR::plot_fn_obj(df_obj = data_obj_try, trend = F, facet_grid = c("indicator", "unit"),
            manual_ylab = "Species", manual_title = NA)
```

# 7. Save plot

This will save the plot to the correct folder.
Adjust height & width using (height=, width=, unit="in") if needed.

```{r}

ggsave(plot_otter,filename=plot_filename_otter, width = 6.5, height = 6, units = "in")

ggsave(plot_try,filename=plot_filename_try, width = 6.5, height = 6, units = "in")
```
