---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-04-29

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(tidyverse)
library(vegan)
library(rioja)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "reef_fish_abundance"

csv_filename<-here(paste0("data/formatted/formatted_csvs/", root_name, "_formatted.csv")) 
object_filename<-here(paste0("data/formatted/final_objects/", root_name, "_object.rds"))
plot_filename<-here(paste0("figures/plots/", root_name, "_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
dat = read.csv(here("data/unformatted/All_Species_Combined_Table_SeaFiSh_2023.csv"))
```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).
Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.


First start with a heatmap to show relative change across all species at the same time.
```{r}
head(dat)

# scale data (z-score) for comparison
plot_dat = dat %>% 
  group_by(Species, Region) %>% 
  mutate(scaled_abundance = scale(StdWtYhat)) %>% 
  ungroup()

# plot heatmap
ggplot(plot_dat, aes(x = YEAR, y = Species, fill = scaled_abundance)) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "RdYlBu", 
                     direction = -1,     
                     name = "Relative\nAbundance\n(Z-Score)",
                     na.value = "grey90") +
  scale_x_continuous(breaks = seq(min(plot_dat$YEAR), max(plot_dat$YEAR), by = 2)) +
  labs(title = "Reef Fish Relative Abundance", subtitle = "Standardized indices scaled by species to show annual deviations", x = "Year", y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~Region)

```

Identify points where community composition changed the most using clustering.
```{r}
# prepare a matrix (years as rows, species as columns)
# Gulf region first
matrix_dat <- dat %>%
  filter(Region == "Gulf") %>%
  select(YEAR, Species, StdWtYhat) %>%
  pivot_wider(names_from = Species, values_from = StdWtYhat) %>%
  drop_na() %>%  # PCA/Clustering requires no missing values
  arrange(YEAR) %>%
  column_to_rownames("YEAR")
  
# Run Constrained Clustering (CONISS)
dist_mat <- vegdist(matrix_dat[,-1], method = "euclidean")
clust_out <- chclust(dist_mat, method = "coniss")
plot(clust_out, hang = -1, main = "Constrained Cluster (CONISS)")
bstick(clust_out)

# Cut the tree into exactly 2 groups (as confirmed by Broken Stick plot)
regimes <- cutree(clust_out, k = 2)
shift_year <- names(regimes)[which(diff(regimes) != 0) + 1]
print(paste("The mathematical regime shift occurred in:", shift_year))



# East
matrix_dat <- dat %>%
  filter(Region == "East") %>%
  select(YEAR, Species, StdWtYhat) %>%
  pivot_wider(names_from = Species, values_from = StdWtYhat) %>%
  drop_na() %>%  # PCA/Clustering requires no missing values
  arrange(YEAR) %>%
  column_to_rownames("YEAR")
  
# Run Constrained Clustering (CONISS)
dist_mat <- vegdist(matrix_dat[,-1], method = "euclidean")
clust_out <- chclust(dist_mat, method = "coniss")
plot(clust_out, hang = -1, main = "Constrained Cluster (CONISS)")
bs = bstick(clust_out)

# Cut the tree into exactly 2 groups (as confirmed by Broken Stick plot)
regimes <- cutree(clust_out, k = 2)
shift_year <- names(regimes)[which(diff(regimes) != 0) + 1]
print(paste("The mathematical regime shift occurred in:", shift_year))
```


```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = c("")
unit_names = c("")
extent_names = c("")

formatted_data = IEAnalyzeR::convert_cleaned_data(your_data, indicator_names, unit_names, extent_names)
```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.
```{r}
write.csv(formatted_data, file = csv_filename, row.names = F)
```


# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.
```{r}
data_obj<-IEAnalyzeR::data_prep(csv_filename)
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.
```{r}
saveRDS(data_obj, file = object_filename)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE)
```

# 7. Save plot

This will save the plot to the correct folder.
Adjust height & width using (height=, width=, unit="in") if needed.

```{r}
ggsave(filename = plot_filename)
```
