---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-04-29. Data were sent to us by Kate Overly

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(tidyverse)
library(vegan)
library(rioja)
library(plotTimeSeries)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "reef_fish_abundance"

csv_filename<-here(paste0("data/formatted/formatted_csvs/", root_name, "_formatted.csv")) 
object_filename<-here(paste0("data/formatted/final_objects/", root_name, "_object.rds"))
plot_filename<-here(paste0("figures/plots/", root_name, "_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
dat = read.csv(here("data/unformatted/All_Species_Combined_Table_SeaFiSh_2023.csv"))
```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).
Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.


Mandy's traffic light code to view relative trends for all species
```{r}
#create a Species-Region combined column
dat$Sp_Reg = paste0(dat$Species,".",dat$Region) 

matrix_dat <- as.data.frame(dat %>%
  select(YEAR, Sp_Reg, StdWtYhat) %>%
  pivot_wider(names_from = Sp_Reg, values_from = StdWtYhat) %>%
  arrange(YEAR))

YEAR = data.frame(YEAR = seq(1993,2023, 1))
YEAR$YEAR = as.numeric(YEAR$YEAR)
matrix_dat2 <- left_join(YEAR, matrix_dat)

str(matrix_dat2)

#view plot        
par(mar = c(5, 11, 4, 2))
plotTimeSeries::trafficLightPlot(matrix_dat2, noNAs = 0.8, cexlabs = 0.9, splits = 4, method = "nmds")
```

save plot
```{r}
png(filename = plot_filename, 
    width = 10,           
    height = 7,           
    units = "in", 
    res = 300)           
par(mar = c(5, 11, 4, 2))
plotTimeSeries::trafficLightPlot(matrix_dat2, noNAs = 0.8, cexlabs = 0.9, splits = 4, method = "nmds")
dev.off()
```


Identify points where community composition changed the most using clustering.
```{r}
# prepare a matrix (years as rows, species as columns)
matrix_dat <- dat %>%
  select(YEAR, Sp_Reg, StdWtYhat) %>%
  pivot_wider(names_from = Sp_Reg, values_from = StdWtYhat) %>%
  drop_na() %>%  
  arrange(YEAR) %>%
  column_to_rownames("YEAR")
  
# FIX: Use the entire matrix since YEAR is already gone from the columns!
dist_mat <- vegdist(matrix_dat, method = "euclidean")

clust_out <- chclust(dist_mat, method = "coniss")
plot(clust_out, hang = -1, main = "Constrained Cluster (CONISS)")

# This will now have finite values to plot!
bstick(clust_out, ng = 9)

# Cut the tree into exactly 2 groups (as confirmed by Broken Stick plot)
regimes <- cutree(clust_out, k = 2)
shift_year <- names(regimes)[which(diff(regimes) != 0) + 1]
print(paste("The mathematical regime shift occurred in:", shift_year))
```

