---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-05-28

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
devtools::install_git("https://github.com/Gulf-IEA/IEAnalyzeR", branch = "fix-partial_dates_data")

library(IEAnalyzeR)
library(here)
library(ggplot2)

library(ncdf4)
library(dplyr)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "SLR"

csv_filename<-here(paste0("data/formatted/formatted_csvs/", root_name, "_formatted.csv")) 
object_filename<-here(paste0("data/formatted/final_objects/", root_name, "_object.rds"))
plot_filename<-here(paste0("figures/plots/", root_name, "_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

Data found [here](https://www.star.nesdis.noaa.gov/socd/lsa/SeaLevelRise/LSA_SLR_timeseries.php)

```{r}
slr_unformatted<-read.csv("../../../data/unformatted/slr_sla_goa_free_ref_90.csv", skip = 5)
```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).
Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.

The data set has some overlapping time periods during satellite switches. During those periods it is mainly just going bak and forth between satellites, but sometimes there are values from two satellites for the same time period. For those few instances I am averaging the value. 
```{r}

slr_satmerge<-slr_unformatted %>% 
  mutate(
    top_jas1=ifelse(!is.na(TOPEX.Poseidon) & !is.na(Jason.1), (TOPEX.Poseidon+Jason.1)/2,
                    ifelse(!is.na(TOPEX.Poseidon), TOPEX.Poseidon, Jason.1)),
    jas1_jas2=ifelse(!is.na(Jason.1) & !is.na(Jason.2), (Jason.1+Jason.2)/2,
                    ifelse(!is.na(Jason.1), Jason.1, Jason.2)),
    jas2_jas3=ifelse(!is.na(Jason.2) & !is.na(Jason.3), (Jason.2+Jason.3)/2,
                    ifelse(!is.na(Jason.2), Jason.2, Jason.3)),
    jas3_sent=ifelse(!is.na(Jason.3) & !is.na(Sentinel.6MF), (Jason.3+Sentinel.6MF)/2,
                    ifelse(!is.na(Jason.3), Jason.3, Sentinel.6MF)))

slr_all<-slr_satmerge %>%  mutate(
  all_sats=ifelse(!is.na(top_jas1), top_jas1,
                  ifelse(!is.na(jas1_jas2), jas1_jas2,
                         ifelse(!is.na(jas1_jas2), jas1_jas2,
                                ifelse(!is.na(jas2_jas3), jas2_jas3,jas3_sent))))) %>% 
    select(year, slr=all_sats)
```

```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = c("Sea Level Rise")
unit_names = c("Change in mean sea level (mm)")
extent_names = c("Gulf of America")

formatted_data = IEAnalyzeR::convert_cleaned_data(slr_all, indicator_names, unit_names, extent_names)
```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.
```{r}
write.csv(formatted_data, file = csv_filename, row.names = F)
```


# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.
```{r}
data_obj<-IEAnalyzeR::data_prep(df=formatted_data)
```

```{r}
formatted_data
clean_dates_to_partial_year(formatted_data$indicator[3:nrow(formatted_data)])

test<-formatted_data$indicator[3:nrow(formatted_data)]

!all(grepl("^[0-9]{4}$", test) | is.na(test))
```

```{r}
df = formatted_data
trends = T
subind = FALSE
anomaly = NULL
library(lubridate)

poly_fix <- function(x, y, mean) {
        d <- data.frame(x = x, y = y)
        rx <- do.call("rbind", lapply(1:(nrow(d) - 1), function(i) {
            f <- lm(x ~ y, d[i:(i + 1), ])
            if (f$qr$rank < 2) 
                return(NULL)
            r <- predict(f, newdata = data.frame(y = mean))
            if (d[i, ]$x < r & r < d[i + 1, ]$x) 
                return(data.frame(x = r, y = mean))
            else return(NULL)
        }))
        d2 <- cbind(rx, mean, mean)
        names(d2) <- c("year", "value", "min", "max")
        return(d2)
}

    df_nm <- paste0(deparse(substitute(df)))
    if (class(df) == "character") {
        df_nm <- sub("^.*/(.*)\\.[^.]*$", "\\1", df)
        df <- read.csv(df, check.names = F, )
    }
    if (!is.null(anomaly)) {
        df_nm <- paste(df_nm, anomaly, "anom", sep = "_")
    }
    df_list <- list()
    is_row3_numeric <- FALSE
    for (i in 2:ncol(df)) {
        cell_value <- as.character(df[3, i])
        if (cell_value != "" && !is.na(cell_value)) {
            is_row3_numeric <- !is.na(suppressWarnings(as.numeric(cell_value)))
            break
        }
    }
    if (is_row3_numeric) {
        data_start_row <- 3
        df_lab <- rbind(c(colnames(df)), as.character(df[1, ]), 
            as.character(df[2, ]))
    } else {
        data_start_row <- 4
        df_lab <- rbind(as.character(df[1, ]), as.character(df[2, 
            ]), as.character(df[3, ]))
    }
    df_dat <- df[data_start_row:nrow(df), c(1:ncol(df))]
    options(scipen = 999)
    clean_dates_to_partial_year <- function(date_strings) {
        date_formats = c("m-Y", "bY", "Ym", "Yb", "b-y", "y-b")
        parsed_dates = parse_date_time(date_strings, orders = date_formats, 
            quiet = TRUE)
        partial_years = year(parsed_dates) + (month(parsed_dates) - 
            1)/12
        return(partial_years)
    }
    if (!all(grepl("^[0-9]{4}(\\.[0-9]+)?$", df_dat[, 1]) | is.na(df_dat[,1]))) {
        df_dat[, 1] <- clean_dates_to_partial_year(df_dat[, 1])
    }
    if (!is.null(anomaly)) {
        get_month <- function(pyear) {
            round((pyear - floor(pyear)) * 12 + 1)
        }
        if (any(na.omit(as.numeric(df_dat[, 1]))%%1 != 0)) {
            sub_list <- list()
            for (i in 2:ncol(df_dat)) {
                sub_df <- df_dat[, c(1, i)]
                colnames(sub_df) <- c("year", "value")
                sub_df$value <- as.numeric(sub_df$value)
                sub_df$month <- get_month(sub_df$year)
                mon_means_sd <- sub_df %>% group_by(month) %>% 
                  summarise(mon_mean = mean(value, na.rm = T), 
                    mon_sd = sd(value, na.rm = T))
                sub_df <- left_join(sub_df, mon_means_sd, by = "month")
                sub_df <- sub_df %>% mutate(anom_value = value - 
                  mon_mean, sd_anom_value = anom_value/mon_sd)
                sub_df$col_nm <- colnames(df)[i]
                sub_df$id <- i - 1
                if (anomaly == "monthly") {
                  sub_df <- dplyr::select(sub_df, year, value = anom_value, 
                    col_nm, id)
                  sub_list[[i - 1]] <- sub_df
                }
                if (anomaly == "stdmonthly") {
                  sub_df <- dplyr::select(sub_df, year, value = sd_anom_value, 
                    col_nm, id)
                  sub_list[[i - 1]] <- sub_df
                }
            }
            df_dat <- do.call("rbind", sub_list)
            df_dat <- df_dat %>% pivot_wider(id_cols = c("year"), 
                names_from = c("col_nm", "id"), values_from = "value")
            colnames(df_dat) <- sub("_(?!.*_).*", "", colnames(df_dat), 
                perl = TRUE)
        }
    }
    if (ncol(df_dat) < 3) {
        colnames(df_dat) <- c("year", "value")
        df_dat$value <- as.numeric(df_dat$value)
        df_dat$year <- as.numeric(df_dat$year)
        df_dat <- df_dat[!is.na(df_dat$value), ]
    } else {
        sub_list <- list()
        for (i in 2:ncol(df_dat)) {
            sub_df <- df_dat[, c(1, i)]
            ind <- ifelse(subind == "extent", df_lab[3, i], ifelse(subind == 
                "unit", df_lab[2, i], df_lab[1, i]))
            colnames(sub_df) <- c("year", "value")
            sub_df <- as.data.frame(lapply(sub_df, as.numeric))
            sub_df$year <- as.numeric(sub_df$year)
            sub_df$subnm <- paste0(ind)
            sub_df$id <- i - 1
            sub_df <- sub_df[!is.na(sub_df$value), ]
            sub_list[[i]] <- sub_df
        }
        df_dat <- do.call("rbind", sub_list)
    }
    df_dat <- df_dat %>% arrange(year)
    df_list$data <- df_dat
    if (ncol(df_dat) < 3) {
        ribbon <- df_dat
        mean <- mean(as.numeric(ribbon$value), na.rm = T)
        ribbon$min <- ifelse(ribbon$value >= mean, mean, ribbon$value)
        ribbon$max <- ifelse(ribbon$value >= mean, ribbon$value, 
            mean)
        poly_o <- poly_fix(ribbon$year, ribbon$value, mean)
        ribbon <- rbind(ribbon, poly_o)
        ribbon$mean <- mean
        ribbon <- ribbon[!is.na(ribbon$value), ]
    } else {
        sub_list <- list()
        subs <- unique(df_dat$id)
        subnm_un <- unique(select(df_dat, subnm, id))
        for (i in 1:length(subs)) {
            ribbon <- df_dat[df_dat$id == subs[i], ]
            mean <- mean(as.numeric(ribbon$value), na.rm = T)
            ribbon$min <- ifelse(ribbon$value >= mean, mean, 
                ribbon$value)
            ribbon$max <- ifelse(ribbon$value >= mean, ribbon$value, 
                mean)
            poly_o <- poly_fix(ribbon$year, ribbon$value, mean)
            poly_o$subnm <- ribbon$subnm[1]
            poly_o$id <- ribbon$id[1]
            ribbon <- rbind(ribbon, poly_o)
            ribbon$mean <- mean
            ribbon <- ribbon[!is.na(ribbon$value), ]
            sub_list[[i]] <- ribbon
        }
        ribbon <- do.call("rbind", sub_list)
    }
    df_list$ribbon <- ribbon
    df_list$labs <- df_lab
    if (trends == T) {
        if (ncol(df_dat) < 3) {
            mean <- mean(as.numeric(df_dat$value), na.rm = T)
            sd <- sd(as.numeric(df_dat$value), na.rm = T)
            minyear <- min(df_dat$year)
            maxyear <- max(df_dat$year)
            allminyear <- min(df_dat$year)
            allmaxyear <- max(df_dat$year)
            last5 <- df_dat[df_dat$year > max(df_dat$year) - 
                5, ]
            last5_mean <- mean(last5$value)
            mean_tr <- if_else(last5_mean > mean + sd, "circle_plus", 
                if_else(last5_mean < mean - sd, "circle_minus", 
                  "circle_fill"))
            mean_sym <- if_else(last5_mean > mean + sd, "+", 
                if_else(last5_mean < mean - sd, "-", "●"))
            mean_word <- if_else(last5_mean > mean + sd, "greater", 
                if_else(last5_mean < mean - sd, "below", "within"))
            lmout <- summary(lm(last5$value ~ last5$year))
            last5_slope <- coef(lmout)[2, 1] * 5
            slope_tr <- if_else(last5_slope > sd, "arrow_up", 
                if_else(last5_slope < c(-sd), "arrow_down", "arrow_leftright"))
            slope_sym <- if_else(last5_slope > sd, "↑", if_else(last5_slope < 
                c(-sd), "↓", "→"))
            slope_word <- if_else(last5_slope > sd, "an increasing", 
                if_else(last5_slope < c(-sd), "a decreasing", 
                  "a stable"))
            mean_img <- system.file("images/trend_symb", paste0(mean_tr, 
                ".png"), package = "IEAnalyzeR")
            slope_img <- system.file("images/trend_symb", paste0(slope_tr, 
                ".png"), package = "IEAnalyzeR")
            mid_y <- (diff(range(na.omit(df_dat)$value))/2) + 
                min(na.omit(df_dat)$value)
            min_y <- min(na.omit(df_dat)$value)
            max_y <- max(na.omit(df_dat)$value)
            vals <- data.frame(mean = mean, sd = sd, minyear = minyear, 
                maxyear = maxyear, allminyear = allminyear, allmaxyear = allmaxyear, 
                mean_tr = mean_tr, mean_img = mean_img, slope_tr = slope_tr, 
                slope_img = slope_img, mean_sym = mean_sym, slope_sym = slope_sym, 
                mean_word = mean_word, slope_word = slope_word, 
                mid_y = mid_y, min_y = min_y, max_y = max_y, 
                df_nm = df_nm)
            if (nrow(last5) < 5) {
                trend_vars <- c("mean_tr", "mean_img", "slope_tr", 
                  "slope_img", "mean_sym", "slope_sym", "mean_word", 
                  "slope_word")
                vals[trend_vars] <- NA
            }
            vals
        }
        else {
            sub_list <- list()
            subnm_un <- unique(select(df_dat, subnm, id))
            subs <- unique(df_dat$id)
            for (i in 1:length(subs)) {
                sub_df <- df_dat[df_dat$id == subs[i], ]
                minyear <- min(na.omit(sub_df)$year)
                maxyear <- max(na.omit(sub_df)$year)
                allminyear <- min(df_dat$year)
                allmaxyear <- max(df_dat$year)
                mean <- mean(as.numeric(sub_df$value), na.rm = T)
                sd <- sd(as.numeric(sub_df$value), na.rm = T)
                last5 <- sub_df[sub_df$year > max(sub_df$year) - 
                  5, ]
                last5_mean <- mean(last5$value)
                mean_tr <- if_else(last5_mean > mean + sd, "circle_plus", 
                  if_else(last5_mean < mean - sd, "circle_minus", 
                    "circle_fill"))
                mean_sym <- if_else(last5_mean > mean + sd, "+", 
                  if_else(last5_mean < mean - sd, "-", "●"))
                mean_word <- if_else(last5_mean > mean + sd, 
                  "greater", if_else(last5_mean < mean - sd, 
                    "below", "within"))
                lmout <- summary(lm(last5$value ~ last5$year))
                last5_slope <- coef(lmout)[2, 1] * 5
                slope_tr <- if_else(last5_slope > sd, "arrow_up", 
                  if_else(last5_slope < c(-sd), "arrow_down", 
                    "arrow_leftright"))
                slope_sym <- if_else(last5_slope > sd, "↑", 
                  if_else(last5_slope < c(-sd), "↓", "→"))
                slope_word <- if_else(last5_slope > sd, "an increasing", 
                  if_else(last5_slope < c(-sd), "a decreasing", 
                    "a stable"))
                mean_img <- system.file("images/trend_symb", 
                  paste0(mean_tr, ".png"), package = "IEAnalyzeR")
                slope_img <- system.file("images/trend_symb", 
                  paste0(slope_tr, ".png"), package = "IEAnalyzeR")
                mid_y <- (diff(range(na.omit(sub_df)$value))/2) + 
                  min(na.omit(sub_df)$value)
                min_y <- min(na.omit(sub_df)$value)
                max_y <- max(na.omit(sub_df)$value)
                vals <- data.frame(allminyear = allminyear, allmaxyear = allmaxyear, 
                  minyear = minyear, maxyear = maxyear, mean = mean, 
                  sd = sd, mean_tr = mean_tr, mean_img = mean_img, 
                  slope_tr = slope_tr, slope_img = slope_img, 
                  mean_sym = mean_sym, slope_sym = slope_sym, 
                  mean_word = mean_word, slope_word = slope_word, 
                  subnm = subnm_un[i, 1], id = unique(sub_df$id), 
                  mid_y = mid_y, min_y = min_y, max_y = max_y, 
                  df_nm = df_nm)
                if (nrow(last5) < 5) {
                  trend_vars <- c("mean_tr", "mean_img", "slope_tr", 
                    "slope_img", "mean_sym", "slope_sym", "mean_word", 
                    "slope_word")
                  vals[trend_vars] <- NA
                }
                sub_list[[i]] <- vals
            }
            vals <- do.call("rbind", sub_list)
        }
        df_list$vals <- vals
    }
    else {
        if (ncol(df_dat) < 3) {
            mean <- mean(as.numeric(df_dat$value), na.rm = T)
            sd <- sd(as.numeric(df_dat$value), na.rm = T)
            minyear <- min(df_dat$year)
            maxyear <- max(df_dat$year)
            allminyear <- min(df_dat$year)
            allmaxyear <- max(df_dat$year)
            vals <- data.frame(mean = mean, sd = sd, minyear = minyear, 
                maxyear = maxyear, allminyear = allminyear, allmaxyear = allmaxyear, 
                df_nm = df_nm)
            vals
        }
        else {
            sub_list <- list()
            subnm_un <- unique(select(df_dat, subnm, id))
            subs <- unique(df_dat$id)
            for (i in 1:length(subs)) {
                sub_df <- df_dat[df_dat$id == subs[i], ]
                minyear <- min(na.omit(sub_df)$year)
                maxyear <- max(na.omit(sub_df)$year)
                allminyear <- min(df_dat$year)
                allmaxyear <- max(df_dat$year)
                mean <- mean(as.numeric(sub_df$value), na.rm = T)
                sd <- sd(as.numeric(sub_df$value), na.rm = T)
                vals <- data.frame(allminyear = allminyear, allmaxyear = allmaxyear, 
                  minyear = minyear, maxyear = maxyear, mean = mean, 
                  sd = sd, subnm = subnm_un[i, 1], id = unique(sub_df$id), 
                  df_nm = df_nm)
                sub_list[[i]] <- vals
            }
            vals <- do.call("rbind", sub_list)
        }
        df_list$vals <- vals
    }
    df_list
    
clean_dates_to_partial_year

parse
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.
```{r}
saveRDS(data_obj, file = object_filename)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE)
```

# 7. Save plot

This will save the plot to the correct folder.
Adjust height & width using (height=, width=, unit="in") if needed.

```{r}
ggsave(filename = plot_filename)
```
