---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-07-09

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)

library(terra)
library(tidyterra)
library(tigris)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "SLR_city"

csv_filename<-here(paste0("data/formatted/formatted_csvs/", root_name, "_formatted.csv")) 
object_filename<-here(paste0("data/formatted/final_objects/", root_name, "_object.rds"))
plot_filename<-here(paste0("figures/plots/", root_name, "_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
pt_slr<-read.csv("../../../data/unformatted/techrpt86_PaP_of_HTFlooding.csv", skip=10)
```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).
Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.


Data also here
https://tidesandcurrents.noaa.gov/sltrends/mslUSTrendsTable.html
```{r}
#Load and rename data
slr_txt<-read.delim("https://tidesandcurrents.noaa.gov/sltrends/data/USStationsLinearSeaLevelTrends.txt", sep = "",
                    col.names = c("stat_id","stat_nm", "yr_first", "yr_last", "yr_range", "per_comp", "trend_mmyr","ci_mmyr", "trend_ftcent", "ci_ftcent", "lat", "long"))

#Format data and cut to gulf cities
slr_txt<-tidyr::separate(slr_txt, stat_nm, sep = ", ", into = c("city", "state"), remove = F)
gulf_states<-c("TX", "LA", "MS", "AL", "FL")
slr_gulf<-slr_txt[slr_txt$state %in% gulf_states,]
atl_rem<-c("Fernandina Beach","Mayport","Port Canaveral", "Lake Worth Pier", "Virginia Key")
slr_gulf<-slr_gulf[!slr_gulf$city %in% atl_rem, ]

```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

- Not standard timeseries data
```{r, eval=F}
write.csv(formatted_data, file = csv_filename, row.names = F)
```


# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

- Not standard timeseries data
```{r, eval=F}
data_obj<-IEAnalyzeR::data_prep(csv_filename)
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

- Not standard timeseries data
```{r, eval=F}
saveRDS(data_obj, file = object_filename)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
#Get gulf states map outline
gulf_states<-states(cb = T, resolution = "5m")
state_nms<-c("Texas", "Louisiana", "Mississippi", "Alabama", "Georgia","Florida")
gulf_states<-gulf_states[gulf_states$NAME %in% state_nms,]
gulf_states<-gulf_states$geometry
gulf_states<-as_spatvector(gulf_states)

#Cut to partial state outlines
new_bb<-ext(-99.3,-80, 22.87229, 32)
new_bb<-vect(new_bb, crs(gulf_states))
gulf_out<-crop(gulf_states, new_bb)

#Plot
ggplot(slr_gulf, aes(x=long, y=lat))+
  geom_spatvector(data = gulf_out, fill="grey80", color="white", inherit.aes=F)+
  geom_point(aes(fill=trend_mmyr), shape=21, color="white", stroke=0.5, size=1.75)+
  scale_fill_viridis_c()+
  scale_x_continuous(expand = c(0,0))+
  scale_y_continuous(expand = c(0,0))+
  labs(fill="RSL Trend\n(mm/yr)")+
  theme_void()+
  theme(legend.position = c(0.5, 0.25),
                      legend.direction = "horizontal",
                      legend.text = element_text(size=7),
                      legend.title = element_text(size=8),
                      legend.key.height = unit(0.5, "cm"),
                      plot.margin = margin(3,3,3,3,"pt"))
```

# 7. Save plot

This will save the plot to the correct folder.
Adjust height & width using (height=, width=, unit="in") if needed.

```{r}
ggsave(filename = plot_filename, height = 3, width = 5.5, units = "in", bg= "white")
```
