---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-06-15

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)

library(terra)
library(tidyterra)
library(tigris)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "SLR_spatial"

# csv_filename<-here(paste0("data/formatted/formatted_csvs/", root_name, "_formatted.csv"))
# object_filename<-here(paste0("data/formatted/final_objects/", root_name, "_object.rds"))
plot_filename<-here(paste0("figures/plots/", root_name, "_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
grid<-terra::rast("../../../data/unformatted/slr_map_ref.nc")
```

# 2. Format data
- NOT A TIMESERIES, ALTERNATE CHUNK WRITTEN

```{r create gulf eez shape}

#Cut grid to gulf eez
eez<-vect("../../../data/intermediate/gulf_eez/eez.shx")

#Bounding Box
gulf_box<-ext(-100, -80.25, 20, 32)
gulf_box<-vect(gulf_box, crs(eez))

#Shape Point DF
gulf<-data.frame(lon= c(-100,-100, -82.97, -80.445,-80.37222,-80.37222),
           lat= c(20,32, 32, 25.292, 25.174,20))
gulf<-vect(gulf,geom=c("lon", "lat"), crs(eez))
gulf<-convHull(gulf)

#Crop gulf EEZ
gulf_eez<-crop(eez, gulf)

grid_gulf<-crop(grid, gulf_eez)
grid_gulf<-mask(grid_gulf, gulf_eez)

```



# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
#cut state outline to gulf states
gulf_states<-states(cb = T, resolution = "5m")
state_nms<-c("Texas", "Louisiana", "Mississippi", "Alabama", "Georgia","Florida")
gulf_states<-gulf_states[gulf_states$NAME %in% state_nms,]
gulf_states<-gulf_states$geometry

#cut gulf states to not full state
new_bb<-ext(-99.3,-80, 22.87229, 32)
new_bb<-vect(new_bb, crs(eez))

gulf_states<-as_spatvector(gulf_states)
gulf_out<-crop(gulf_states, new_bb)
# plot(gulf_out)

ggplot()+
  geom_spatraster(data=grid_gulf)+
  geom_spatvector(data = gulf_out, fill="grey80", color="white")+
  # geom_spatvector(data = gulf_eez, fill=NA, color="black")+
  scale_fill_viridis_c(na.value = "transparent", option = "D")+
  scale_x_continuous(expand = c(0,0))+
  scale_y_continuous(expand = c(0,0))+
  labs(fill="Sea Level\nTrends (mm/yr)")+
  theme_void()+ theme(legend.position = c(0.25, 0.1),
                      legend.direction = "horizontal",
                      plot.margin = margin(3,3,3,3,"pt"))
```

# 7. Save plot

This will save the plot to the correct folder.
Adjust height & width using (height=, width=, unit="in") if needed.

```{r}
ggsave(filename = plot_filename, height = 3, width = 5.5, units = "in")
```

