---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-04-09

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "AMO"

csv_filename<-here(paste0("data/formatted/formatted_csvs/", root_name, "_formatted.csv")) 
object_filename<-here(paste0("data/formatted/final_objects/", root_name, "_object.rds"))
plot_filename<-here(paste0("figures/plots/", root_name, "_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

Info regarding metric: https://psl.noaa.gov/data/timeseries/AMO/

Notes from code of NCAR calculation: 
; Generate the Atlantic Multi-decadal Oscillation (AMO) Index
;
; The AMO is a coherent pattern of variability in basin-wide 
; North Atlantic SSTs with a period of about 60-80 yrs.
;
; To isolate the actual AMO the background global annual
; sst means or anomalies are removed.
```{r}
#AMO data hosted online in text form
url<-"https://www.ncei.noaa.gov/pub/data/cmb/ersst/v5/index/ersst.v5.amo.dat"

#Read in data using the online file
raw<-read.table(url, header = T,skip = 1)

#Combine year month
amo<-as.data.frame(cbind(year=paste(raw$Year, raw$month, sep="-"), SSTA=raw$SSTA))


#Numeric
amo$SSTA<-as.numeric(amo$SSTA)
```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).
Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.

```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = c("Atlantic Multidecadal Oscillation")
unit_names = c("SST Anomaly")
extent_names = c("Northern Atlantic")

formatted_data = IEAnalyzeR::convert_cleaned_data(amo, indicator_names, unit_names, extent_names)
```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.
```{r}
write.csv(formatted_data, file = csv_filename, row.names = F)
```


# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.
```{r}
data_obj<-IEAnalyzeR::data_prep(csv_filename)
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.
```{r}
saveRDS(data_obj, file = object_filename)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, lwd = 0.5, xbreaks_by = 10)
```

# 7. Save plot

This will save the plot to the correct folder.
Adjust height & width using (height=, width=, unit="in") if needed.

```{r}
ggsave(filename = plot_filename, height = 3, units = "in")
```
