---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-07-21 by Carissa Gervasi.
Required to have a login to Oracle to pull data, but non-confidential summaries are stored on Github.
The documentation for analyzing commercial landings data can be found in the Gulf IEA google drive folder for the 2025 Gulf ESR.

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(ROracle)
library(keyring)
library(tidyverse)
library(scales)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "com_landings"

csv_filename1<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_assessed_formatted.csv")) 
object_filename1<-here(paste0("data/formatted/final_objects/", root_name, "all_assessed_object.rds"))
plot_filename1<-here(paste0("figures/plots/", root_name, "all_assessed_plot.png"))

csv_filename2<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_unassessed_formatted.csv")) 
object_filename2<-here(paste0("data/formatted/final_objects/", root_name, "all_unassessed_object.rds"))
plot_filename2<-here(paste0("figures/plots/", root_name, "all_unassessed_plot.png"))

```

# 1. Read Data - SKIP THIS SECTION UNLESS YOU NEED TO PULL DATA FROM ORACLE AGAIN

Pull data from its source: Manual data: data/unformatted data.
Automated data: Add script for data call (API, package, etc.).
Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo.
If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
Sys.setenv(TZ = "UTC", ORA_SDTZ = "UTC")

# Define your username
my_username <- "CGERVASI"

# Connect using stored keyring password
con <- dbConnect(dbDriver("Oracle"), 
                 username = my_username,
                 password = key_get("SECPR_Oracle", my_username),
                 dbname = "SECPR")

# Run quick test query
#test_query <- "SELECT * FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV WHERE ROWNUM <= 1000"
#df <- dbGetQuery(con, test_query)

# SQL Query filtered at database level
query <- "SELECT LAND_YEAR, SPECIES_ITIS, COMMON_NAME, LIVE_POUNDS, DOLLARS, DEALER_ID, FLAG_FMP_SPP 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
          AND LAND_YEAR >= 1972
          AND REGION IN ('GOM', 'GOM INTL')"
```

```{r}
# Fetch data
landings_raw <- dbGetQuery(con, query)
```

```{r}
#Always close the connection when you're done
dbDisconnect(con)
```

Now we need to subset the giant dataset to what we need and save the intermediate files.
First, we will pull out landings and revenue data for the 15 species that have stock assessments.

```{r}
# Filter & Prepare Clean Working Data
# ------------------------------------------------------------------------------
clean_data <- landings_raw %>%
  filter(!is.na(LAND_YEAR)) %>%
  mutate(fmp_status = if_else(FLAG_FMP_SPP == 1, "Managed", "Unmanaged"))

```

```{r}
# Separate out assessed species
# ------------------------------------------------------------------------------

target_species_exact <- c(
  "cobia", 
  "grouper, gag", 
  "snapper, gray", 
  "triggerfish, gray", 
  "amberjack, greater", 
  "hogfish", 
  "mackerel, king", 
  "snapper, mutton", 
  "grouper, red", 
  "snapper, red", 
  "scamp", 
  "mackerel, spanish", 
  "snapper, vermilion", 
  "grouper, yellowedge", 
  "snapper, yellowtail"
)

# Summarize 15 core assessed species
summary_target_15 <- clean_data %>%
  filter(
    tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  group_by(LAND_YEAR, COMMON_NAME) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

summary_target_15 = summary_target_15 %>% 
  filter(is_confidential != TRUE)  # Remove any confidential rows with fewer than 3 dealers

# Verify which unique species names matched from your database
cat("Matched species in dataset:\n")
print(unique(summary_target_15$COMMON_NAME))

# Save target list summary
write_csv(summary_target_15, here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
```

```{r}
# Now the unassessed species

# Keywords for reef fish families 
reef_family_keywords <- "PORGY|GRUNT|SNAPPER|GROUPER|SEA BASS|SEABASS|TRIGGERFISH|JACK|WRASSE|PARROTFISH|TILEFISH|HOGFISH|HIND|CONEY|GRAYSBY|TOMTATE|PINFISH"

# Filter and assign Guilds to Unassessed Reef Fish
unassessed_reef_fish <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>% 
  group_by(LAND_YEAR, COMMON_NAME, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_fish2 = unassessed_reef_fish %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef fish not confidential
write_csv(unassessed_reef_fish2, here("data/intermediate/gulf_unassessed_reef_fish_1972_present.csv"))

print(unique(unassessed_reef_fish$COMMON_NAME))

#  [1] "GROUPER, GOLIATH"      "GROUPER, WARSAW"       "JACK, CREVALLE"        "SNAPPER, LANE"        
#  [5] "TILEFISH, GOLDEN"      "GROUPER, SNOWY"        "GROUPER, YELLOWMOUTH"  "SNAPPER, BLACK"       
#  [9] "SNAPPER, BLACKFIN"     "SNAPPER, QUEEN"        "SNAPPER, SILK"         "GROUPER, BLACK"       
# [13] "HIND, SPECKLED"        "TUNA, SKIPJACK"        "GROUPER, MARBLED"      "GROUPER, NASSAU"      
# [17] "GROUPER, YELLOWFIN"    "HIND, RED"             "HIND, ROCK"            "SNAPPER, MAHOGANY"    
# [21] "TILEFISH, BLUELINE"    "PORGY, RED"            "GRAYSBY"               "GROUPER, MISTY"       
# [25] "PORGY, WHITEBONE"      "AMBERJACK, LESSER"     "GROUPER, CONEY"        "GRUNT, FRENCH"        
# [29] "HOGFISH, SPANISH"      "JACK, ALMACO"          "JACKKNIFE-FISH"        "PARROTFISH, BLUE"     
# [33] "PARROTFISH, MIDNIGHT"  "PARROTFISH, RAINBOW"   "PARROTFISH, REDBAND"   "PARROTFISH, STRIPED"  
# [37] "SNAPPER, CUBERA"       "SNAPPER, DOG"          "SNAPPER, GLASSEYE"     "TOMTATE"              
# [41] "TRIGGERFISH, QUEEN"    "WRASSE, CLOWN"         "WRASSE, CREOLE"        "WRASSE, YELLOWHEAD"   
# [45] "GROUPER, TIGER"        "JACK, BAR"             "JACK, BLACK"           "PARROTFISH, PRINCESS" 
# [49] "PARROTFISH, QUEEN"     "PARROTFISH, STOPLIGHT" "TILEFISH, GOLDFACE"    "TILEFISH, SAND"       
# [53] "JACK, HORSE-EYE"       "JACK, YELLOW"          "PORGY, JOLTHEAD"       "PORGY, KNOBBED"       
# [57] "TILEFISH, BLACKLINE"   "PORGY, LONGSPINE"      "TRIGGERFISH, OCEAN"    "GRUNT, WHITE" 


unassessed_reef_summary <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>%
  group_by(LAND_YEAR, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_summary = unassessed_reef_summary %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef summary
write_csv(unassessed_reef_summary, here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
```

```{r}
# OPTIONAL: Free up RAM by deleting the 40-million row raw dataset
rm(landings_raw, clean_data)
gc()
```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).

Start from here if you are just using the intermediate datasets.

```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish = dat_assessed %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)



dat_unassessed_summary = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed_summary)

dat_unassessed_summary$total_mil_lbs = dat_unassessed_summary$total_live_lbs / 1e6
dat_unassessed_summary$total_mil_dollars = dat_unassessed_summary$total_dollars / 1e6

unassessed_finfish_all = dat_unassessed_summary %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
```

Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function.
Replace "your_data" in the code below with whatever your dataframe is called.

```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
extent_names = c("Assessed finfish", "Assessed finfish")

formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(assessed_finfish, indicator_names, unit_names, extent_names)


#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
extent_names = rep("Unassessed finfish", 2)

formatted_unassessed_all = IEAnalyzeR::convert_cleaned_data(unassessed_finfish_all, indicator_names, unit_names, extent_names)
```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

```{r}
write.csv(formatted_assessed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unassessed_all, file = csv_filename2, row.names = F)
```

# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Assessed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Unassessed Finfish")
ggsave(filename = plot_filename2)
```

------------------------------------------------------------------------

Now that we have overall trends, we want to break down by species.
Starting with the assessed species.

```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish_species = dat_assessed %>% 
  group_by(LAND_YEAR, COMMON_NAME) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, COMMON_NAME, total_mil_lbs, total_mil_dollars)


for (i in unique(assessed_finfish_species$COMMON_NAME)) {
  
  species_data = assessed_finfish_species %>% 
    filter(COMMON_NAME == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_assessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```



Now let's do the same for the unassessed species.

```{r}
dat_unassessed = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed)

dat_unassessed$total_mil_lbs = dat_unassessed$total_live_lbs / 1e6
dat_unassessed$total_mil_dollars = dat_unassessed$total_dollars / 1e6

unassessed_finfish_species = dat_unassessed %>% 
  group_by(LAND_YEAR, guild) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, guild, total_mil_lbs, total_mil_dollars)


for (i in unique(unassessed_finfish_species$guild)) {
  
  species_data = unassessed_finfish_species %>% 
    filter(guild == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_unassessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_unassessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```


