---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-07-21 by Carissa Gervasi.
Required to have a login to Oracle to pull data, but non-confidential summaries are stored on Github.
The documentation for analyzing commercial landings data can be found in the Gulf IEA google drive folder for the 2025 Gulf ESR.

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(ROracle)
library(keyring)
library(tidyverse)
library(scales)
library(stringr)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "com_landings"

csv_filename1<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_assessed_formatted.csv")) 
object_filename1<-here(paste0("data/formatted/final_objects/", root_name, "all_assessed_object.rds"))
plot_filename1<-here(paste0("figures/plots/", root_name, "all_assessed_plot.png"))

csv_filename2<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_unassessed_formatted.csv")) 
object_filename2<-here(paste0("data/formatted/final_objects/", root_name, "all_unassessed_object.rds"))
plot_filename2<-here(paste0("figures/plots/", root_name, "all_unassessed_plot.png"))

```

# 1. Read Data - SKIP THIS SECTION UNLESS YOU NEED TO PULL DATA FROM ORACLE AGAIN

Pull data from its source: Manual data: data/unformatted data.
Automated data: Add script for data call (API, package, etc.).
Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo.
If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
Sys.setenv(TZ = "UTC", ORA_SDTZ = "UTC")

# Define your username
my_username <- "CGERVASI"

# Connect using stored keyring password
con <- dbConnect(dbDriver("Oracle"), 
                 username = my_username,
                 password = key_get("SECPR_Oracle", my_username),
                 dbname = "SECPR")

# Run quick test query
#test_query <- "SELECT * FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV WHERE ROWNUM <= 1000"
#df <- dbGetQuery(con, test_query)

# SQL Query filtered at database level
query <- "SELECT LAND_YEAR, SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP, SUM(LIVE_POUNDS) AS TOTAL_LIVE_POUNDS, SUM(DOLLARS) AS TOTAL_DOLLARS, COUNT(*) AS RECORD_COUNT 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
            AND LAND_YEAR >= 1972
            AND REGION IN ('GOM', 'GOM INTL')
          GROUP BY
            LAND_YEAR, SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP
          ORDER BY
            LAND_YEAR ASC, TOTAL_LIVE_POUNDS DESC"
```

```{r}
# Fetch data
landings_raw <- dbGetQuery(con, query)
```

Quick check to ensure none of the data is confidential

```{r}
query2 <- "SELECT LAND_YEAR, COMMON_NAME, COUNT(DISTINCT DEALER_ID) AS UNIQUE_DEALERS, SUM(LIVE_POUNDS) AS TOTAL_LIVE_LBS 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
            AND LAND_YEAR >= 1972
            AND REGION IN ('GOM', 'GOM INTL')
          GROUP BY LAND_YEAR, COMMON_NAME
          HAVING COUNT(DISTINCT DEALER_ID) < 3 AND COUNT(DISTINCT DEALER_ID) > 0"
```

```{r}
conf_check <- dbGetQuery(con, query2)
unique(conf_check$COMMON_NAME)
```

There are 4,867 confidential entries.
They all look like broad categories or freshwater fish.
We can safely ignore them.

```{r}
#Always close the connection when you're done
dbDisconnect(con)
```

Now we need to subset the giant dataset to what we need and save the intermediate files.
First, some checks to see how the broad and unidentified categories are distributed.

```{r}
unique(landings_raw$COMMON_NAME)

# Refined filter to target ONLY true aggregate/unclassified categories
aggregate_taxa <- landings_raw %>%
  filter(
    # 1. Matches explicit unclassified keywords
    str_detect(COMMON_NAME, regex("UNCLASSIFIED|UNIDENTIFIED|OTHER|GENUS|FAMILY|NOS|SPEC", ignore_case = TRUE)) |
    # 2. Exact single-word family names (e.g., "GROUPERS" or "SNAPPER" alone)
    str_to_upper(trimws(COMMON_NAME)) %in% c("GROUPER", "GROUPERS", "SNAPPER", "SNAPPERS", 
                                            "JACK", "JACKS", "MACKEREL", "MACKERELS", 
                                            "TILEFISH", "TILEFISHES", "GRUNT", "GRUNTS",
                                            "FISH", "FINFISH", "FISHES")
  ) %>%
  group_by(SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    MIN_YEAR = min(LAND_YEAR),
    MAX_YEAR = max(LAND_YEAR),
    .groups = "drop"
  ) %>%
  arrange(desc(TOTAL_LIVE_LBS))

# Print the true broad/aggregate categories
print(head(aggregate_taxa, 30))
```


Before we move on, let's save a working file so we don't have to pull from Oracle again.
We will use the conf_check dataset to ensure none of the confidential data are included.

```{r}
# 1. Exclude the confidential strata identified in conf_check
landings_github <- landings_raw %>%
  # Anti-join matches on year, species name, and total live pounds to remove confidential rows
  anti_join(
    conf_check, 
    by = c("LAND_YEAR", "COMMON_NAME")
  )

# 2. Save compressed CSV file for GitHub repository
write_csv(landings_github, here("data/intermediate/gom_commercial_landings_1972_2025_clean.csv"))

cat("Confidential rows successfully removed!\n")
cat("Original rows:", nrow(landings_raw), "\n")
cat("Cleaned dataset rows:", nrow(landings_github), "\n")
```





# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).

Start from here if you are just using the saved dataset.

```{r}
# Load data
landings_github = read_csv(here("data/intermediate/gom_commercial_landings_1972_2025_clean.csv"))
```

Ok so the broad categories are not included as managed species according to the FLAG_FMP_SPP column. If we ignore them, we might ignore a lot of landings.

This plot is to check if we really need to include those uncategorized broad groups. How much of the total landings do they account for?

```{r}
# Gulf Council Finfish FMP Broad Groups
gulf_council_aggregates <- c(
  # Reef Fish FMP Aggregates
  "GROUPERS", "UNCLASSIFIED GROUPERS", "SNAPPER", "SNAPPERS", "UNCLASSIFIED SNAPPERS",
  "TILEFISHES", "PORGIES", "TRIGGERFISH", "TRIGGERFISHES", "GRUNT", "GRUNTS", 
  "WRASSE", "WRASSES", "HOGFISH", "AMBERJACKS", "JACKS", "RUNNERS",
  # Coastal Migratory Pelagics (CMP) FMP Aggregates
  "MACKERELS", "UNCLASSIFIED MACKERELS", "POMPANOS",
  # Red Drum FMP / Sciaenids Aggregates
  "DRUMS", "SEATROUTS"
)

# Optional: Highly Migratory Species (HMS) - Managed Federally / Secretarial, not directly GMFMC
hms_aggregates <- c("SHARK", "SHARK, UNIDENTIFIED", "SHARKS", "TUNAS", "BILLFISHES", "SWORDFISH")

# Categorize into Council-relevant groupings
plot1_council_df <- landings_github %>%
  filter(LAND_YEAR >= 1972) %>%
  mutate(
    # Ensure uppercase clean matching
    NAME_UPPER = str_to_upper(trimws(COMMON_NAME)),
    
    Category = case_when(
      FLAG_FMP_SPP == 1 ~ "Species-Specific Managed Landings",
      NAME_UPPER %in% gulf_council_aggregates ~ "Unclassified Managed Family Groups (GMFMC)",
      NAME_UPPER %in% hms_aggregates ~ "HMS / Pelagic Family Groups",
      TRUE ~ "Other / Unmanaged"
    )
  ) %>%
  filter(Category != "Other / Unmanaged") %>%
  group_by(LAND_YEAR, Category) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

# Plot for ESR Presentation
ggplot(plot1_council_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6, fill = Category)) +
  geom_area(alpha = 0.85, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c(
    "Species-Specific Managed Landings" = "#2c7fb8", 
    "Unclassified Managed Family Groups (GMFMC)" = "#fe9929",
    "HMS / Pelagic Family Groups" = "#7fcdbb"
  )) +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Gulf of Mexico Commercial Managed Finfish Extraction (1972–Present)",
    subtitle = "Ecosystem Status Report: Evaluated by management flagging and aggregate reporting categories",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    fill = "Reporting & Management Strata",
    caption = "Data Source: NOAA SEFSC Commercial Landings Extraction MV. Excludes shellfish/shrimp."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


YAY the uncategorized group are a small blip. We can safely move forward with just plotting the landings for all species categorized as FLAG_FMP_SPP == 1 (Managed). 


```{r}
# Filter & Prepare Clean Working Data
# ------------------------------------------------------------------------------
clean_data <- landings_github %>%
  filter(!is.na(LAND_YEAR)) %>%
  mutate(fmp_status = if_else(FLAG_FMP_SPP == 1, "Managed", "Unmanaged"))

managed_data <- clean_data %>%
  filter(fmp_status == "Managed")

unique(managed_data$COMMON_NAME)

```


clean the dataset to only keep species within Gulf Council Finfish FMPs.
```{r}
# Strictly verified GMFMC-managed finfish species
gmfmc_finfish_official <- c(
  # Reef Fish FMP - Snappers
  "SNAPPER, RED", "SNAPPER, VERMILION", "SNAPPER, GRAY", "SNAPPER, LANE", 
  "SNAPPER, MUTTON", "SNAPPER, YELLOWTAIL", "SNAPPER, CUBERA", "SNAPPER, SILK", 
  "SNAPPER, BLACKFIN", "SNAPPER, QUEEN", "WENCHMAN",
  
  # Reef Fish FMP - Groupers
  "GROUPER, GAG", "GROUPER, RED", "SCAMP", "GROUPER, YELLOWEDGE", 
  "GROUPER, BLACK", "GROUPER, YELLOWFIN", "GROUPER, YELLOWMOUTH", 
  "GROUPER, SNOWY", "GROUPER, WARSAW", "GROUPER, GOLIATH", "GROUPER, NASSAU",
  
  # Reef Fish FMP - Tilefishes
  "TILEFISH, GOLDEN", "TILEFISH, BLUELINE", "TILEFISH, GOLDFACE",
  
  # Reef Fish FMP - Jacks
  "AMBERJACK, GREATER", "AMBERJACK, LESSER", "JACK, ALMACO",
  
  # Reef Fish FMP - Others
  "TRIGGERFISH, GRAY", "HOGFISH",
  
  # Coastal Migratory Pelagics (CMP) FMP
  "MACKEREL, KING", "MACKEREL, SPANISH", "COBIA",
  
  # Red Drum FMP
  "DRUM, RED"
)

# Filter landings_github to strictly official GMFMC finfish
gmfmc_landings_df <- landings_github %>%
  filter(
    LAND_YEAR >= 1972,
    COMMON_NAME %in% gmfmc_finfish_official
  ) %>%
  group_by(LAND_YEAR, COMMON_NAME) %>%
  summarise(
    TOTAL_LIVE_POUNDS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
# Summarize annual landings using ONLY clean GMFMC finfish species
plot1_clean_df <- gmfmc_landings_df %>%
  group_by(LAND_YEAR) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

# Generate plot
ggplot(plot1_clean_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6)) +
  geom_area(fill = "#2c7fb8", alpha = 0.75, color = "#023858", linewidth = 0.8) +
  geom_point(color = "#023858", size = 1.5) +
  scale_y_continuous(labels = comma_format(), limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Gulf of Mexico Commercial Managed Finfish Extraction (1972–Present)",
    subtitle = "Ecosystem Status Report: Filtered to GMFMC Managed Finfish (Excludes Shellfish & Non-GMFMC Taxa)",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    caption = "Data Source: NOAA SEFSC Commercial Landings Extraction MV."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey30", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Now let's just compare this to the core priority species managed by the council.

```{r}

unique(gmfmc_landings_df$COMMON_NAME)

# Vector of your core priority species
core_priority_species <- c(
  # Drums / CMP / Pelagics
  "DRUM, RED", "COBIA", "MACKEREL, SPANISH", "MACKEREL, KING", "HOGFISH",
  
  # Snappers
  "SNAPPER, RED", "SNAPPER, CUBERA", "SNAPPER, LANE", "SNAPPER, GRAY", 
  "SNAPPER, MUTTON", "SNAPPER, YELLOWTAIL", "SNAPPER, VERMILION",
  
  # Mid-water / Deepwater Snappers (Mid-water snapper complex components)
  "SNAPPER, SILK", "SNAPPER, BLACKFIN", "SNAPPER, QUEEN", "WENCHMAN",
  
  # Groupers
  "GROUPER, RED", "GROUPER, GAG", "GROUPER, YELLOWEDGE", "SCAMP",
  
  # Jacks (Jack complex components)
  "AMBERJACK, GREATER", "AMBERJACK, LESSER", "JACK, ALMACO",
  
  # Tilefishes (Tilefish complex components)
  "TILEFISH, GOLDEN", "TILEFISH, BLUELINE", "TILEFISH, GOLDFACE",
  
  # Triggerfish
  "TRIGGERFISH, GRAY"
)

dat_assessed <- landings_github %>%
  filter(
    COMMON_NAME %in% core_priority_species)

dat_assessed$total_mil_lbs = dat_assessed$TOTAL_LIVE_POUNDS / 1e6
dat_assessed$total_mil_dollars = dat_assessed$TOTAL_DOLLARS / 1e6

assessed_finfish = dat_assessed %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)

write_csv(assessed_finfish, here("data/intermediate/gulf_core_priority_species_landings_1972_present.csv"))
```


This is just a check to make sure we are not missing anything important in the core species list.
```{r}
# 1. Prepare combined comparative dataset
comparison_df <- gmfmc_landings_df %>%
  mutate(
    Series = if_else(
      COMMON_NAME %in% core_priority_species,
      "Core Priority Species Subset",
      "Other Managed GMFMC Finfish"
    )
  ) %>%
  group_by(LAND_YEAR, Series) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Generate Comparative Stacked Area Plot
ggplot(comparison_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6, fill = Series)) +
  geom_area(alpha = 0.8, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c(
    "Core Priority Species Subset" = "#08519c", 
    "Other Managed GMFMC Finfish" = "#9ecae1"
  )) +
  scale_y_continuous(labels = comma_format(), limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Gulf of Mexico Managed Commercial Finfish Extraction (1972–Present)",
    subtitle = "Comparing Total GMFMC Managed Finfish Landings vs. Core Priority Species Subset",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    fill = "Species Selection Strategy",
    caption = "Data Source: NOAA SEFSC Commercial Landings Extraction MV. Excludes shellfish and non-GMFMC taxa."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey30", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function.
Replace "your_data" in the code below with whatever your dataframe is called.

```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
extent_names = c("Assessed finfish", "Assessed finfish")

formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(assessed_finfish, indicator_names, unit_names, extent_names)

```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

```{r}
write.csv(formatted_assessed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unassessed_all, file = csv_filename2, row.names = F)
```

# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Managed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Unassessed Finfish")
ggsave(filename = plot_filename2)
```


































```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish = dat_assessed %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)



dat_unassessed_summary = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed_summary)

dat_unassessed_summary$total_mil_lbs = dat_unassessed_summary$total_live_lbs / 1e6
dat_unassessed_summary$total_mil_dollars = dat_unassessed_summary$total_dollars / 1e6

unassessed_finfish_all = dat_unassessed_summary %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
```















































```{r}
# Now the unassessed species

# Keywords for reef fish families 
reef_family_keywords <- "PORGY|GRUNT|SNAPPER|GROUPER|SEA BASS|SEABASS|TRIGGERFISH|JACK|WRASSE|PARROTFISH|TILEFISH|HOGFISH|HIND|CONEY|GRAYSBY|TOMTATE|PINFISH"

# Filter and assign Guilds to Unassessed Reef Fish
unassessed_reef_fish <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>% 
  group_by(LAND_YEAR, COMMON_NAME, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_fish2 = unassessed_reef_fish %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef fish not confidential
write_csv(unassessed_reef_fish2, here("data/intermediate/gulf_unassessed_reef_fish_1972_present.csv"))

print(unique(unassessed_reef_fish$COMMON_NAME))


unassessed_reef_summary <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>%
  group_by(LAND_YEAR, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_summary = unassessed_reef_summary %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef summary
write_csv(unassessed_reef_summary, here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
```

```{r}
# OPTIONAL: Free up RAM by deleting the 40-million row raw dataset
rm(landings_raw, clean_data)
gc()
```





# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

```{r}
write.csv(formatted_assessed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unassessed_all, file = csv_filename2, row.names = F)
```

# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Assessed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Unassessed Finfish")
ggsave(filename = plot_filename2)
```

------------------------------------------------------------------------

Now that we have overall trends, we want to break down by species.
Starting with the assessed species.

```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish_species = dat_assessed %>% 
  group_by(LAND_YEAR, COMMON_NAME) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, COMMON_NAME, total_mil_lbs, total_mil_dollars)


for (i in unique(assessed_finfish_species$COMMON_NAME)) {
  
  species_data = assessed_finfish_species %>% 
    filter(COMMON_NAME == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_assessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```

Now let's do the same for the unassessed species.

```{r}
dat_unassessed = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed)

dat_unassessed$total_mil_lbs = dat_unassessed$total_live_lbs / 1e6
dat_unassessed$total_mil_dollars = dat_unassessed$total_dollars / 1e6

unassessed_finfish_species = dat_unassessed %>% 
  group_by(LAND_YEAR, guild) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, guild, total_mil_lbs, total_mil_dollars)


for (i in unique(unassessed_finfish_species$guild)) {
  
  species_data = unassessed_finfish_species %>% 
    filter(guild == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_unassessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_unassessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```
