---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-07-21 by Carissa Gervasi.
Required to have a login to Oracle to pull data, but non-confidential summaries are stored on Github.
The documentation for analyzing commercial landings data can be found in the Gulf IEA google drive folder for the 2025 Gulf ESR.

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(ROracle)
library(keyring)
library(tidyverse)
library(scales)
library(stringr)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "com_landings"

csv_filename1<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_assessed_formatted.csv")) 
object_filename1<-here(paste0("data/formatted/final_objects/", root_name, "all_assessed_object.rds"))
plot_filename1<-here(paste0("figures/plots/", root_name, "all_assessed_plot.png"))

csv_filename2<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_unassessed_formatted.csv")) 
object_filename2<-here(paste0("data/formatted/final_objects/", root_name, "all_unassessed_object.rds"))
plot_filename2<-here(paste0("figures/plots/", root_name, "all_unassessed_plot.png"))

```

# 1. Read Data - SKIP THIS SECTION UNLESS YOU NEED TO PULL DATA FROM ORACLE AGAIN

Pull data from its source: Manual data: data/unformatted data.
Automated data: Add script for data call (API, package, etc.).
Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo.
If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
Sys.setenv(TZ = "UTC", ORA_SDTZ = "UTC")

# Define your username
my_username <- "CGERVASI"

# Connect using stored keyring password
con <- dbConnect(dbDriver("Oracle"), 
                 username = my_username,
                 password = key_get("SECPR_Oracle", my_username),
                 dbname = "SECPR")

# Run quick test query
#test_query <- "SELECT * FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV WHERE ROWNUM <= 1000"
#df <- dbGetQuery(con, test_query)

# SQL Query filtered at database level
query <- "SELECT LAND_YEAR, SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP, SUM(LIVE_POUNDS) AS TOTAL_LIVE_POUNDS, SUM(DOLLARS) AS TOTAL_DOLLARS, COUNT(*) AS RECORD_COUNT 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
            AND LAND_YEAR >= 1972
            AND REGION IN ('GOM', 'GOM INTL')
          GROUP BY
            LAND_YEAR, SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP
          ORDER BY
            LAND_YEAR ASC, TOTAL_LIVE_POUNDS DESC"
```

```{r}
# Fetch data
landings_raw <- dbGetQuery(con, query)
```

Quick check to ensure none of the data is confidential

```{r}
query2 <- "SELECT LAND_YEAR, COMMON_NAME, COUNT(DISTINCT DEALER_ID) AS UNIQUE_DEALERS, SUM(LIVE_POUNDS) AS TOTAL_LIVE_LBS 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
            AND LAND_YEAR >= 1972
            AND REGION IN ('GOM', 'GOM INTL')
          GROUP BY LAND_YEAR, COMMON_NAME
          HAVING COUNT(DISTINCT DEALER_ID) < 3 AND COUNT(DISTINCT DEALER_ID) > 0"
```

```{r}
conf_check <- dbGetQuery(con, query2)
unique(conf_check$COMMON_NAME)
```

There are 4,867 confidential entries.
They all look like broad categories or freshwater fish.
We can safely ignore them.

```{r}
#Always close the connection when you're done
dbDisconnect(con)
```

Now we need to subset the giant dataset to what we need and save the intermediate files.
First, some checks to see how the broad and unidentified categories are distributed.

```{r}
unique(landings_raw$COMMON_NAME)

# Refined filter to target ONLY true aggregate/unclassified categories
aggregate_taxa <- landings_raw %>%
  filter(
    # 1. Matches explicit unclassified keywords
    str_detect(COMMON_NAME, regex("UNCLASSIFIED|UNIDENTIFIED|OTHER|GENUS|FAMILY|NOS|SPEC", ignore_case = TRUE)) |
    # 2. Exact single-word family names (e.g., "GROUPERS" or "SNAPPER" alone)
    str_to_upper(trimws(COMMON_NAME)) %in% c("GROUPER", "GROUPERS", "SNAPPER", "SNAPPERS", 
                                            "JACK", "JACKS", "MACKEREL", "MACKERELS", 
                                            "TILEFISH", "TILEFISHES", "GRUNT", "GRUNTS",
                                            "FISH", "FINFISH", "FISHES")
  ) %>%
  group_by(SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    MIN_YEAR = min(LAND_YEAR),
    MAX_YEAR = max(LAND_YEAR),
    .groups = "drop"
  ) %>%
  arrange(desc(TOTAL_LIVE_LBS))

# Print the true broad/aggregate categories
print(head(aggregate_taxa, 30))
```


Before we move on, let's save a working file so we don't have to pull from Oracle again.
We will use the conf_check dataset to ensure none of the confidential data are included.

```{r}
# 1. Exclude the confidential strata identified in conf_check
landings_github <- landings_raw %>%
  # Anti-join matches on year, species name, and total live pounds to remove confidential rows
  anti_join(
    conf_check, 
    by = c("LAND_YEAR", "COMMON_NAME")
  )

# 2. Save compressed CSV file for GitHub repository
write_csv(landings_github, here("data/intermediate/gom_commercial_landings_1972_2025_clean.csv"))

cat("Confidential rows successfully removed!\n")
cat("Original rows:", nrow(landings_raw), "\n")
cat("Cleaned dataset rows:", nrow(landings_github), "\n")
```
###############################################################################################

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).

Start from here if you are just using the saved dataset.

```{r}
# Load data
landings_github = read_csv(here("data/intermediate/gom_commercial_landings_1972_2025_clean.csv"))
```

```{r}
# Filter & Prepare Clean Working Data
clean_data <- landings_github %>%
  filter(!is.na(LAND_YEAR)) %>%
  mutate(fmp_status = if_else(FLAG_FMP_SPP == 1, "Managed", "Unmanaged"))

managed_data <- clean_data %>%
  filter(fmp_status == "Managed")

unique(managed_data$COMMON_NAME)
```


clean the dataset to only keep species within Gulf Council Finfish FMPs, the Coastal Migratory Pelagics (CMP) FMP, and the red drum FMP. 
https://www.fisheries.noaa.gov/species/federally-managed-gulf-america-reef-fish
https://www.fisheries.noaa.gov/management-plan/gulf-and-south-atlantic-coastal-migratory-pelagic-fishery-management-plan 
https://www.fisheries.noaa.gov/management-plan/gulf-red-drum-fishery-management-plan 
```{r}
# Strictly verified GMFMC-managed finfish species
gmfmc_finfish_official <- c(
  # Reef Fish FMP - Snappers
  "SNAPPER, RED", "SNAPPER, VERMILION", "SNAPPER, GRAY", "SNAPPER, LANE", 
  "SNAPPER, MUTTON", "SNAPPER, YELLOWTAIL", "SNAPPER, CUBERA", "SNAPPER, SILK", 
  "SNAPPER, BLACKFIN", "SNAPPER, QUEEN", "WENCHMAN",
  
  # Reef Fish FMP - Groupers
  "GROUPER, GAG", "GROUPER, RED", "SCAMP", "GROUPER, YELLOWEDGE", 
  "GROUPER, BLACK", "GROUPER, YELLOWFIN", "GROUPER, YELLOWMOUTH", 
  "GROUPER, SNOWY", "GROUPER, WARSAW", "GROUPER, GOLIATH", "GROUPER, NASSAU", "HIND, SPECKLED",
  
  # Reef Fish FMP - Tilefishes
  "TILEFISH, GOLDEN", "TILEFISH, BLUELINE", "TILEFISH, GOLDFACE",
  
  # Reef Fish FMP - Jacks
  "AMBERJACK, GREATER", "AMBERJACK, LESSER", "JACK, ALMACO", "RUDDERFISH, BANDED",
  
  # Reef Fish FMP - Others
  "TRIGGERFISH, GRAY", "HOGFISH", 
  
  # Coastal Migratory Pelagics (CMP) FMP
  "MACKEREL, KING", "MACKEREL, SPANISH", "COBIA",
  
  # Red Drum FMP
  "DRUM, RED"
)

# Filter landings to strictly official GMFMC finfish

# 1. Load the GDP deflator dataset and aggregate to annual average deflators
deflator_annual <- read_csv(here("data/intermediate/gdp_deflator_interpolated.csv")) %>%
  mutate(LAND_YEAR = as.numeric(format(as.Date(date), "%Y"))) %>%
  group_by(LAND_YEAR) %>%
  summarise(annual_deflator = mean(gdp_deflator_dollar, na.rm = TRUE), .groups = "drop")

# Define target baseline deflator (2024 dollars)
target_deflator_2024 <- deflator_annual %>%
  filter(LAND_YEAR == 2024) %>%
  pull(annual_deflator)

dat_managed <- clean_data %>%
  filter(COMMON_NAME %in% gmfmc_finfish_official) %>%
  left_join(deflator_annual, by = "LAND_YEAR") %>%
  mutate(
    # Convert nominal dollars to constant 2024 dollars
    TOTAL_DOLLARS_2024 = TOTAL_DOLLARS * (target_deflator_2024 / annual_deflator),
    total_mil_lbs = TOTAL_LIVE_POUNDS / 1e6,
    total_mil_dollars_2024 = TOTAL_DOLLARS_2024 / 1e6
  )

managed_finfish <- dat_managed %>% 
  group_by(LAND_YEAR) %>%
  summarise(
    total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), 
    total_mil_dollars = sum(total_mil_dollars_2024, na.rm = TRUE), 
    .groups = "drop"
  ) %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)

write_csv(dat_managed, here("data/intermediate/managed_species_landings_by_species.csv"))
write_csv(managed_finfish, here("data/intermediate/gulf_managed_species_com_landings_1972_present.csv"))
```



Ok so the broad categories are not included as managed species according to the FLAG_FMP_SPP column. If we ignore them, we might ignore a lot of landings.

This plot is to check if we really need to include those uncategorized broad groups. How much of the total landings do they account for?

```{r}
# First need to try to figure out how many broad groups exist. Theoretically all the broad groups should have single words for the species name. 
broad_groups = clean_data %>% 
  filter(
    # Matches "unclassified" or "unknown" (case-insensitive)
    str_detect(COMMON_NAME, regex("unclassified|unknown", ignore_case = TRUE)) |
    # Matches single words (no spaces)
    !str_detect(COMMON_NAME, "\\s")) %>% 
  distinct(COMMON_NAME) %>% 
  arrange(COMMON_NAME)

write_csv(broad_groups, here("data/intermediate/gulf_broad_groups_commercial_list.csv"))

# Now let's pull out all the broad groups that may contain our federally managed species. 
gulf_council_aggregates <- c(
  "AMBERJACKS",
  "DRUMS",
  "GROUPERS",
  "JACKS",
  "MACKERELS",
  "SNAPPERS",
  "TILEFISHES",
  "TRIGGERFISHES",
  "MACKERELS, KING AND CERO"
)

#check
test = clean_data %>% 
  filter(COMMON_NAME %in% gulf_council_aggregates) %>% 
  distinct(COMMON_NAME)


# See how much of the landings are in these broad groups
plot1_council_df <- clean_data %>%
  mutate(
    Category = case_when(
      COMMON_NAME %in% gmfmc_finfish_official ~ "Species-Specific Managed Landings",
      COMMON_NAME %in% gulf_council_aggregates ~ "Unclassified Managed Family Groups",
      TRUE ~ "Other / Unmanaged"
    )
  ) %>%
  filter(Category != "Other / Unmanaged") %>%
  group_by(LAND_YEAR, Category) %>%
  summarise(
    TOTAL_LIVE_POUNDS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
ggplot(plot1_council_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_POUNDS / 1e6, fill = Category)) +
  geom_area(alpha = 0.85, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c(
    "Species-Specific Managed Landings" = "#2c7fb8", 
    "Unclassified Managed Family Groups" = "#fe9929"
  )) +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Gulf Commercial Managed Finfish Extraction (1972–Present)",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    fill = "Reporting & Management Strata"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


Ok so there is a fairly substantial number of landings earlier in the time series for those broad groups. Options:

1. Ignore the broad groups (potentially underestimate landings early in the time series)
2. Cut the time series to 1986, the first full year that the FL trip ticket program was active and species were being reported more consistently.
3. Add the broad category data to the species specific data, possibly overestimating landings
4. Add uncertainty to the graph using upper and lower bounds, with the upper bound being the species-specific landings plus broad group landings and the lower bound being just the species specific landings. 





### Unmanaged species

Let's go back to the clean_data and filter for the unmanaged species we want to plot. First some diagnostics.

```{r}
# Define exclusions (Managed GMFMC species + Shellfish + Broad Aggregates)
shellfish_and_broad <- c(
  "SHRIMP, NORTHERN BROWN", "SHRIMP, NORTHERN WHITE", "SHRIMP, NORTHERN PINK",
  "SHRIMP, ROYAL RED", "SHRIMP, ROCK", "LOBSTER, CARIBBEAN SPINY", "CRAB, GOLDEN DEEPSEA",
  "CRAB, BLUE", "CRAB, STONE", "OYSTER, EASTERN", "SNAIL, OTHER", "STARS, OTHER SEA"
)

# Filter for strictly unmanaged finfish taxa
unmanaged_diagnostics <- clean_data %>%
  filter(
    !COMMON_NAME %in% gmfmc_finfish_official,
    !COMMON_NAME %in% shellfish_and_broad,
    !str_detect(COMMON_NAME, regex("UNCLASSIFIED|UNIDENTIFIED|OTHER|GENUS|FAMILY|NOS|SPEC", ignore_case = TRUE))
  ) %>%
  group_by(SPECIES_ITIS, COMMON_NAME) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    TOTAL_NOMINAL_DOLLARS = sum(TOTAL_DOLLARS, na.rm = TRUE),
    YEARS_PRESENT = n_distinct(LAND_YEAR),
    .groups = "drop"
  ) %>%
  arrange(desc(TOTAL_LIVE_LBS))

# View top 30 unmanaged species by volume
print(head(unmanaged_diagnostics, 30))
```

Menhaden clearly rule the world and there are 529 other species in the dataset...Menhaden clearly have to come out.

Let's run a Pareto distribution & sensitivity analysis on the rest of the species. Calculates the cumulative percentage of total unmanaged marine finfish volume accounted for by each species, ranks them from highest to lowest, and tests if the time series trend changes when using top 5, top 10, top 15, or top 25 species. Trying to remove species was like playing constant whack a mole. Random things just kept popping up. So a better approach was to make a whitelist of known major commercial marine finfish that are not managed. 

```{r}
unmanaged_all = clean_data %>% 
  filter(COMMON_NAME %in% unmanaged_diagnostics$COMMON_NAME, COMMON_NAME != "MENHADENS")

# Cumulative Landings Pareto Analysis
pareto_df <- unmanaged_all %>%
  group_by(COMMON_NAME) %>%
  summarise(TOTAL_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TOTAL_LBS)) %>%
  mutate(
    RANK = row_number(),
    CUM_LBS = cumsum(TOTAL_LBS),
    CUM_PCT = (CUM_LBS / sum(TOTAL_LBS)) * 100
  )

write_csv(pareto_df, here("data/intermediate/unmanaged__com_species_pareto_analysis.csv"))

# We need to remove more. SOOOO MANY RANDOM SPECIES WE DON'T CARE ABOUT

# Define Regex Exclusion Patterns

# Non-finfish / Invertebrates (Shrimp, Crabs, Oysters, Lobsters, Shellfish, Jellyfish, Corals, etc.)
non_finfish_pattern <- "SHRIMP|CRAB|OYSTER|LOBSTER|CRAYFISH|PROCAMBARUS|CLAM|SCALLOP|OCTOPUS|SQUID|JELLYFISH|SPONGE|DOLLAR|TURTLE|TURBINIDAE|CONCH|CORAL|SEAWEED|ANIMALIA|BRYOPSIDALES|FORAMS|NUDIBRANCHS|CRUSTACEANS|SQUIDS|OCTOPUSES|SPONGES|CLAMS|SCALLOPS|HERMIT|STARFISH|SNAIL|ALGAE|WHELK|ANEMONE|NASSARIUS|CERITHIDEA|TEGULA|LIBINIA|MITHRACULUS|HIPPOSPONGIA|CHITONS|CORALLIMORPHARIA|CORALLINACEAE|DICOTYLEDONS|ULOTRICHALES|NASSA|STARS,|TURBAN"

# Freshwater / Estuarine Non-Target Species (Catfish, Buffalo, Carp, Gar, Bowfin, etc.)
freshwater_pattern <- "CATFISH|BUFFALO|CARP|BOWFIN|GAR|TILAPIA|SUNFISH|LARGEMOUTH|DRUM, FRESHWATER|SHAD, GIZZARD|BULLHEAD|PICKEREL|SUCKER|PADDLEFISH|SHAD, AMERICAN"

# HMS / Pelagic Sharks & Tunas / Industrial Menhaden
hms_menhaden_pattern <- "MENHADEN|TUNA|SWORDFISH|SHARK|MARLIN|SAILFISH|SPEARFISH"

# Generic Broad Family Aggregates (Catch-all dealer codes)
aggregate_pattern <- "^OSTEICHTHYES$|^GROUPERS$|^SNAPPERS$|^MACKERELS$|^TILEFISHES$|^JACKS$|^TRIGGERFISHES$|^AMBERJACKS$|^DRUMS$|^FISHES$|MACKERELS, KING AND CERO"

# Combine all patterns into one regular expression
all_exclusions <- paste(non_finfish_pattern, freshwater_pattern, hms_menhaden_pattern, aggregate_pattern, sep = "|")

# 3. Filter and Recalculate True Clean Marine Finfish Pareto Rankings
pareto_df2 <- pareto_df %>%
  # Filter out non-target entries
  filter(!str_detect(COMMON_NAME, regex(all_exclusions, ignore_case = TRUE))) %>%
  # Sort by total pounds descending
  arrange(desc(TOTAL_LBS)) %>%
  # Recalculate Rank, Cumulative Pounds, and Cumulative Percentage
  mutate(
    RANK = row_number(),
    CUM_LBS = cumsum(TOTAL_LBS),
    CUM_PCT = (CUM_LBS / sum(TOTAL_LBS)) * 100
  )


# Sensitivity Analysis: Plot trajectories using Top 5, Top 10, Top 15, vs Top 25
top10_spp  <- pareto_df2 %>% filter(RANK <= 10) %>% pull(COMMON_NAME)
top20_spp <- pareto_df2 %>% filter(RANK <= 20) %>% pull(COMMON_NAME)
top30_spp <- pareto_df2 %>% filter(RANK <= 30) %>% pull(COMMON_NAME)
top40_spp <- pareto_df2 %>% filter(RANK <= 40) %>% pull(COMMON_NAME)

sensitivity_df <- unmanaged_all %>%
  filter(COMMON_NAME %in% top40_spp) %>%
  mutate(
    Cutoff = case_when(
      COMMON_NAME %in% top10_spp ~ "Top 10 Species",
      COMMON_NAME %in% top20_spp ~ "Top 20 Species",
      COMMON_NAME %in% top30_spp ~ "Top 30 Species",
      TRUE ~ "Top 40 Species"
    )
  ) %>%
  group_by(LAND_YEAR, Cutoff) %>%
  summarise(TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE), .groups = "drop")

# Plot sensitivity curves
ggplot(sensitivity_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6, color = Cutoff)) +
  geom_line(linewidth = 1.1) +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Sensitivity of Unmanaged Finfish Trajectory to Taxonomic Scope",
    subtitle = "Testing cumulative volume thresholds (Top 10 vs 20 vs 30 vs 40 species)",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    color = "Inclusion Threshold"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```

Ok now we have a better handle on which marine/coastal finfish are the major commercially harvested unmanaged species. The top 25 species make up 95% of the landings. So let's just plot those. 

```{r}
# Explicit Whitelists: Major Unmanaged Commercial Marine Finfish in the Gulf (top 95% of landings for unmanaged finfish)
# Define sub-categories
top_unmanaged_food_fish <- c(
  "MULLET, STRIPED", "DRUM, BLACK", "LADYFISH", "SHEEPSHEAD", 
  "CROAKER, ATLANTIC", "SEATROUT, SPOTTED", "JACK, CREVALLE", 
  "FLATFISHES", "RUNNER, BLUE", "SEATROUT, SAND", "POMPANO, FLORIDA", 
  "KINGFISHES", "BLUEFISH", "MULLET, WHITE", "DOLPHIN", "DOLPHINFISH"
)

top_unmanaged_bait_forage <- c(
  "HERRING, ATLANTIC THREAD", "SARDINE, SPANISH", "HERRINGS", 
  "SCAD, ROUND", "BUTTERFISH,GULF", "GRUNTS", "PORGIES", "SCADS", "MOJARRAS"
)
```

Now we can get the unmanaged data!

```{r}
# Filter landings to both food and bait lists & adjust for inflation
dat_unmanaged <- clean_data %>%
  filter(
    COMMON_NAME %in% c(top_unmanaged_food_fish, top_unmanaged_bait_forage)
  ) %>%
  left_join(deflator_annual, by = "LAND_YEAR") %>%
  mutate(
    TOTAL_DOLLARS_2024 = TOTAL_DOLLARS * (target_deflator_2024 / annual_deflator),
    total_mil_lbs = TOTAL_LIVE_POUNDS / 1e6,
    total_mil_dollars_2024 = TOTAL_DOLLARS_2024 / 1e6,
    
    # Assign species category for conditional grouping
    Species_Type = case_when(
      COMMON_NAME %in% top_unmanaged_food_fish   ~ "Food",
      COMMON_NAME %in% top_unmanaged_bait_forage ~ "Bait"
    )
  )

# 2. Summarize annual totals into 4 distinct columns
unmanaged_summary <- dat_unmanaged %>% 
  group_by(LAND_YEAR) %>%
  summarise(
    food_mil_lbs     = sum(total_mil_lbs[Species_Type == "Food"], na.rm = TRUE),
    food_mil_dollars = sum(total_mil_dollars_2024[Species_Type == "Food"], na.rm = TRUE),
    bait_mil_lbs     = sum(total_mil_lbs[Species_Type == "Bait"], na.rm = TRUE),
    bait_mil_dollars = sum(total_mil_dollars_2024[Species_Type == "Bait"], na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  select(LAND_YEAR, food_mil_lbs, food_mil_dollars, bait_mil_lbs, bait_mil_dollars)

# Save intermediate summary CSVs
write_csv(dat_unmanaged, here("data/intermediate/unmanaged_species_landings_by_species.csv"))
write_csv(unmanaged_summary, here("data/intermediate/gulf_unmanaged_species_landings_1972_present.csv"))
```



Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function.
Replace "your_data" in the code below with whatever your dataframe is called.

```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Real value (2024 $ million)")
extent_names = c("Assessed finfish", "Assessed finfish")

formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(assessed_finfish, indicator_names, unit_names, extent_names)

#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 4)
unit_names = c("Whole weight (million lbs)", "Real value (2024 $ million)", "Whole weight (million lbs)", "Real value (2024 $ million)")
extent_names = c("Unmanaged food fish", "Unmanaged food fish", "Unmanaged bait/forage fish", "Unmanaged bait/forage fish")

formatted_unassessed_all = IEAnalyzeR::convert_cleaned_data(unmanaged_summary, indicator_names, unit_names, extent_names)

```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

```{r}
write.csv(formatted_assessed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unassessed_all, file = csv_filename2, row.names = F)
```

# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Managed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Top Unmanaged Finfish", ylab_sublabel = c("extent", "unit"))
ggsave(filename = plot_filename2)
```



































































Ignore everything below this...





```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish = dat_assessed %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)



dat_unassessed_summary = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed_summary)

dat_unassessed_summary$total_mil_lbs = dat_unassessed_summary$total_live_lbs / 1e6
dat_unassessed_summary$total_mil_dollars = dat_unassessed_summary$total_dollars / 1e6

unassessed_finfish_all = dat_unassessed_summary %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
```




```{r}
# Now the unassessed species

# Keywords for reef fish families 
reef_family_keywords <- "PORGY|GRUNT|SNAPPER|GROUPER|SEA BASS|SEABASS|TRIGGERFISH|JACK|WRASSE|PARROTFISH|TILEFISH|HOGFISH|HIND|CONEY|GRAYSBY|TOMTATE|PINFISH"

# Filter and assign Guilds to Unassessed Reef Fish
unassessed_reef_fish <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>% 
  group_by(LAND_YEAR, COMMON_NAME, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_fish2 = unassessed_reef_fish %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef fish not confidential
write_csv(unassessed_reef_fish2, here("data/intermediate/gulf_unassessed_reef_fish_1972_present.csv"))

print(unique(unassessed_reef_fish$COMMON_NAME))


unassessed_reef_summary <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>%
  group_by(LAND_YEAR, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_summary = unassessed_reef_summary %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef summary
write_csv(unassessed_reef_summary, here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
```

```{r}
# OPTIONAL: Free up RAM by deleting the 40-million row raw dataset
rm(landings_raw, clean_data)
gc()
```





# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

```{r}
write.csv(formatted_assessed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unassessed_all, file = csv_filename2, row.names = F)
```

# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Assessed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Unassessed Finfish")
ggsave(filename = plot_filename2)
```

------------------------------------------------------------------------

Now that we have overall trends, we want to break down by species.
Starting with the assessed species.

```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish_species = dat_assessed %>% 
  group_by(LAND_YEAR, COMMON_NAME) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, COMMON_NAME, total_mil_lbs, total_mil_dollars)


for (i in unique(assessed_finfish_species$COMMON_NAME)) {
  
  species_data = assessed_finfish_species %>% 
    filter(COMMON_NAME == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_assessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```

Now let's do the same for the unassessed species.

```{r}
dat_unassessed = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed)

dat_unassessed$total_mil_lbs = dat_unassessed$total_live_lbs / 1e6
dat_unassessed$total_mil_dollars = dat_unassessed$total_dollars / 1e6

unassessed_finfish_species = dat_unassessed %>% 
  group_by(LAND_YEAR, guild) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, guild, total_mil_lbs, total_mil_dollars)


for (i in unique(unassessed_finfish_species$guild)) {
  
  species_data = unassessed_finfish_species %>% 
    filter(guild == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_unassessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_unassessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```
