---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-07-21 by Carissa Gervasi.
Required to have a login to Oracle to pull data, but non-confidential summaries are stored on Github.
The documentation for analyzing commercial landings data can be found in the Gulf IEA google drive folder for the 2025 Gulf ESR.

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(ROracle)
library(keyring)
library(tidyverse)
library(scales)
library(stringr)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "com_landings"

csv_filename1<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_assessed_formatted.csv")) 
object_filename1<-here(paste0("data/formatted/final_objects/", root_name, "all_assessed_object.rds"))
plot_filename1<-here(paste0("figures/plots/", root_name, "all_assessed_plot.png"))

csv_filename2<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_unassessed_formatted.csv")) 
object_filename2<-here(paste0("data/formatted/final_objects/", root_name, "all_unassessed_object.rds"))
plot_filename2<-here(paste0("figures/plots/", root_name, "all_unassessed_plot.png"))

```

# 1. Read Data - SKIP THIS SECTION UNLESS YOU NEED TO PULL DATA FROM ORACLE AGAIN

Pull data from its source: Manual data: data/unformatted data.
Automated data: Add script for data call (API, package, etc.).
Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo.
If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
Sys.setenv(TZ = "UTC", ORA_SDTZ = "UTC")

# Define your username
my_username <- "CGERVASI"

# Connect using stored keyring password
con <- dbConnect(dbDriver("Oracle"), 
                 username = my_username,
                 password = key_get("SECPR_Oracle", my_username),
                 dbname = "SECPR")

# Run quick test query
#test_query <- "SELECT * FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV WHERE ROWNUM <= 1000"
#df <- dbGetQuery(con, test_query)

# SQL Query filtered at database level
query <- "SELECT LAND_YEAR, SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP, SUM(LIVE_POUNDS) AS TOTAL_LIVE_POUNDS, SUM(DOLLARS) AS TOTAL_DOLLARS, COUNT(*) AS RECORD_COUNT 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
            AND LAND_YEAR >= 1972
            AND REGION IN ('GOM', 'GOM INTL')
          GROUP BY
            LAND_YEAR, SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP
          ORDER BY
            LAND_YEAR ASC, TOTAL_LIVE_POUNDS DESC"
```

```{r}
# Fetch data
landings_raw <- dbGetQuery(con, query)
```

Quick check to ensure none of the data is confidential

```{r}
query2 <- "SELECT LAND_YEAR, COMMON_NAME, COUNT(DISTINCT DEALER_ID) AS UNIQUE_DEALERS, SUM(LIVE_POUNDS) AS TOTAL_LIVE_LBS 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
            AND LAND_YEAR >= 1972
            AND REGION IN ('GOM', 'GOM INTL')
          GROUP BY LAND_YEAR, COMMON_NAME
          HAVING COUNT(DISTINCT DEALER_ID) < 3 AND COUNT(DISTINCT DEALER_ID) > 0"
```

```{r}
conf_check <- dbGetQuery(con, query2)
unique(conf_check$COMMON_NAME)
```

There are 4,867 confidential entries.
They all look like broad categories or freshwater fish.
We can safely ignore them.

```{r}
#Always close the connection when you're done
dbDisconnect(con)
```

Now we need to subset the giant dataset to what we need and save the intermediate files.
First, some checks to see how the broad and unidentified categories are distributed.

```{r}
unique(landings_raw$COMMON_NAME)

# Refined filter to target ONLY true aggregate/unclassified categories
aggregate_taxa <- landings_raw %>%
  filter(
    # 1. Matches explicit unclassified keywords
    str_detect(COMMON_NAME, regex("UNCLASSIFIED|UNIDENTIFIED|OTHER|GENUS|FAMILY|NOS|SPEC", ignore_case = TRUE)) |
    # 2. Exact single-word family names (e.g., "GROUPERS" or "SNAPPER" alone)
    str_to_upper(trimws(COMMON_NAME)) %in% c("GROUPER", "GROUPERS", "SNAPPER", "SNAPPERS", 
                                            "JACK", "JACKS", "MACKEREL", "MACKERELS", 
                                            "TILEFISH", "TILEFISHES", "GRUNT", "GRUNTS",
                                            "FISH", "FINFISH", "FISHES")
  ) %>%
  group_by(SPECIES_ITIS, COMMON_NAME, FLAG_FMP_SPP) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    MIN_YEAR = min(LAND_YEAR),
    MAX_YEAR = max(LAND_YEAR),
    .groups = "drop"
  ) %>%
  arrange(desc(TOTAL_LIVE_LBS))

# Print the true broad/aggregate categories
print(head(aggregate_taxa, 30))
```


Before we move on, let's save a working file so we don't have to pull from Oracle again.
We will use the conf_check dataset to ensure none of the confidential data are included.

```{r}
# 1. Exclude the confidential strata identified in conf_check
landings_github <- landings_raw %>%
  # Anti-join matches on year, species name, and total live pounds to remove confidential rows
  anti_join(
    conf_check, 
    by = c("LAND_YEAR", "COMMON_NAME")
  )

# 2. Save compressed CSV file for GitHub repository
write_csv(landings_github, here("data/intermediate/gom_commercial_landings_1972_2025_clean.csv"))

cat("Confidential rows successfully removed!\n")
cat("Original rows:", nrow(landings_raw), "\n")
cat("Cleaned dataset rows:", nrow(landings_github), "\n")
```





# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).

Start from here if you are just using the saved dataset.

```{r}
# Load data
landings_github = read_csv(here("data/intermediate/gom_commercial_landings_1972_2025_clean.csv"))
```

Ok so the broad categories are not included as managed species according to the FLAG_FMP_SPP column. If we ignore them, we might ignore a lot of landings.

This plot is to check if we really need to include those uncategorized broad groups. How much of the total landings do they account for?

```{r}
# Gulf Council Finfish FMP Broad Groups
gulf_council_aggregates <- c(
  # Reef Fish FMP Aggregates
  "GROUPERS", "UNCLASSIFIED GROUPERS", "SNAPPER", "SNAPPERS", "UNCLASSIFIED SNAPPERS",
  "TILEFISHES", "PORGIES", "TRIGGERFISH", "TRIGGERFISHES", "GRUNT", "GRUNTS", 
  "WRASSE", "WRASSES", "HOGFISH", "AMBERJACKS", "JACKS", "RUNNERS",
  # Coastal Migratory Pelagics (CMP) FMP Aggregates
  "MACKERELS", "UNCLASSIFIED MACKERELS", "POMPANOS",
  # Red Drum FMP / Sciaenids Aggregates
  "DRUMS", "SEATROUTS"
)

# Optional: Highly Migratory Species (HMS) - Managed Federally / Secretarial, not directly GMFMC
hms_aggregates <- c("SHARK", "SHARK, UNIDENTIFIED", "SHARKS", "TUNAS", "BILLFISHES", "SWORDFISH")

# Categorize into Council-relevant groupings
plot1_council_df <- landings_github %>%
  filter(LAND_YEAR >= 1972) %>%
  mutate(
    # Ensure uppercase clean matching
    NAME_UPPER = str_to_upper(trimws(COMMON_NAME)),
    
    Category = case_when(
      FLAG_FMP_SPP == 1 ~ "Species-Specific Managed Landings",
      NAME_UPPER %in% gulf_council_aggregates ~ "Unclassified Managed Family Groups (GMFMC)",
      NAME_UPPER %in% hms_aggregates ~ "HMS / Pelagic Family Groups",
      TRUE ~ "Other / Unmanaged"
    )
  ) %>%
  filter(Category != "Other / Unmanaged") %>%
  group_by(LAND_YEAR, Category) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

# Plot for ESR Presentation
ggplot(plot1_council_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6, fill = Category)) +
  geom_area(alpha = 0.85, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c(
    "Species-Specific Managed Landings" = "#2c7fb8", 
    "Unclassified Managed Family Groups (GMFMC)" = "#fe9929",
    "HMS / Pelagic Family Groups" = "#7fcdbb"
  )) +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Gulf of Mexico Commercial Managed Finfish Extraction (1972–Present)",
    subtitle = "Ecosystem Status Report: Evaluated by management flagging and aggregate reporting categories",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    fill = "Reporting & Management Strata",
    caption = "Data Source: NOAA SEFSC Commercial Landings Extraction MV. Excludes shellfish/shrimp."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```


YAY the uncategorized group are a small blip. We can safely move forward with just plotting the landings for all species categorized as FLAG_FMP_SPP == 1 (Managed). 


```{r}
# Filter & Prepare Clean Working Data
# ------------------------------------------------------------------------------
clean_data <- landings_github %>%
  filter(!is.na(LAND_YEAR)) %>%
  mutate(fmp_status = if_else(FLAG_FMP_SPP == 1, "Managed", "Unmanaged"))

managed_data <- clean_data %>%
  filter(fmp_status == "Managed")

unique(managed_data$COMMON_NAME)

```


clean the dataset to only keep species within Gulf Council Finfish FMPs.
```{r}
# Strictly verified GMFMC-managed finfish species
gmfmc_finfish_official <- c(
  # Reef Fish FMP - Snappers
  "SNAPPER, RED", "SNAPPER, VERMILION", "SNAPPER, GRAY", "SNAPPER, LANE", 
  "SNAPPER, MUTTON", "SNAPPER, YELLOWTAIL", "SNAPPER, CUBERA", "SNAPPER, SILK", 
  "SNAPPER, BLACKFIN", "SNAPPER, QUEEN", "WENCHMAN",
  
  # Reef Fish FMP - Groupers
  "GROUPER, GAG", "GROUPER, RED", "SCAMP", "GROUPER, YELLOWEDGE", 
  "GROUPER, BLACK", "GROUPER, YELLOWFIN", "GROUPER, YELLOWMOUTH", 
  "GROUPER, SNOWY", "GROUPER, WARSAW", "GROUPER, GOLIATH", "GROUPER, NASSAU",
  
  # Reef Fish FMP - Tilefishes
  "TILEFISH, GOLDEN", "TILEFISH, BLUELINE", "TILEFISH, GOLDFACE",
  
  # Reef Fish FMP - Jacks
  "AMBERJACK, GREATER", "AMBERJACK, LESSER", "JACK, ALMACO",
  
  # Reef Fish FMP - Others
  "TRIGGERFISH, GRAY", "HOGFISH",
  
  # Coastal Migratory Pelagics (CMP) FMP
  "MACKEREL, KING", "MACKEREL, SPANISH", "COBIA",
  
  # Red Drum FMP
  "DRUM, RED"
)

# Filter landings_github to strictly official GMFMC finfish
gmfmc_landings_df <- landings_github %>%
  filter(
    LAND_YEAR >= 1972,
    COMMON_NAME %in% gmfmc_finfish_official
  ) %>%
  group_by(LAND_YEAR, COMMON_NAME) %>%
  summarise(
    TOTAL_LIVE_POUNDS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

```

```{r}
# Summarize annual landings using ONLY clean GMFMC finfish species
plot1_clean_df <- gmfmc_landings_df %>%
  group_by(LAND_YEAR) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

# Generate plot
ggplot(plot1_clean_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6)) +
  geom_area(fill = "#2c7fb8", alpha = 0.75, color = "#023858", linewidth = 0.8) +
  geom_point(color = "#023858", size = 1.5) +
  scale_y_continuous(labels = comma_format(), limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Gulf of Mexico Commercial Managed Finfish Extraction (1972–Present)",
    subtitle = "Ecosystem Status Report: Filtered to GMFMC Managed Finfish (Excludes Shellfish & Non-GMFMC Taxa)",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    caption = "Data Source: NOAA SEFSC Commercial Landings Extraction MV."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey30", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Now let's just compare this to the core priority species managed by the council.

```{r}

unique(gmfmc_landings_df$COMMON_NAME)

# Vector of your core priority species
core_priority_species <- c(
  # Drums / CMP / Pelagics
  "DRUM, RED", "COBIA", "MACKEREL, SPANISH", "MACKEREL, KING", "HOGFISH",
  
  # Snappers
  "SNAPPER, RED", "SNAPPER, CUBERA", "SNAPPER, LANE", "SNAPPER, GRAY", 
  "SNAPPER, MUTTON", "SNAPPER, YELLOWTAIL", "SNAPPER, VERMILION",
  
  # Mid-water / Deepwater Snappers (Mid-water snapper complex components)
  "SNAPPER, SILK", "SNAPPER, BLACKFIN", "SNAPPER, QUEEN", "WENCHMAN",
  
  # Groupers
  "GROUPER, RED", "GROUPER, GAG", "GROUPER, YELLOWEDGE", "SCAMP",
  
  # Jacks (Jack complex components)
  "AMBERJACK, GREATER", "AMBERJACK, LESSER", "JACK, ALMACO",
  
  # Tilefishes (Tilefish complex components)
  "TILEFISH, GOLDEN", "TILEFISH, BLUELINE", "TILEFISH, GOLDFACE",
  
  # Triggerfish
  "TRIGGERFISH, GRAY"
)

# 1. Load the GDP deflator dataset and aggregate to annual average deflators
deflator_annual <- read_csv(here("data/intermediate/gdp_deflator_interpolated.csv")) %>%
  mutate(LAND_YEAR = as.numeric(format(as.Date(date), "%Y"))) %>%
  group_by(LAND_YEAR) %>%
  summarise(annual_deflator = mean(gdp_deflator_dollar, na.rm = TRUE), .groups = "drop")

# Define target baseline deflator (2024 dollars)
target_deflator_2024 <- deflator_annual %>%
  filter(LAND_YEAR == 2024) %>%
  pull(annual_deflator)

dat_assessed <- landings_github %>%
  filter(COMMON_NAME %in% core_priority_species) %>%
  left_join(deflator_annual, by = "LAND_YEAR") %>%
  mutate(
    # Convert nominal dollars to constant 2024 dollars
    TOTAL_DOLLARS_2024 = TOTAL_DOLLARS * (target_deflator_2024 / annual_deflator),
    total_mil_lbs = TOTAL_LIVE_POUNDS / 1e6,
    total_mil_dollars_2024 = TOTAL_DOLLARS_2024 / 1e6
  )

assessed_finfish <- dat_assessed %>% 
  group_by(LAND_YEAR) %>%
  summarise(
    total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), 
    total_mil_dollars = sum(total_mil_dollars_2024, na.rm = TRUE), 
    .groups = "drop"
  ) %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)

write_csv(dat_assessed, here("data/intermediate/managed_species_landings_by_species.csv"))
write_csv(assessed_finfish, here("data/intermediate/gulf_core_priority_species_landings_1972_present.csv"))
```


This is just a check to make sure we are not missing anything important in the core species list.
```{r}
# 1. Prepare combined comparative dataset
comparison_df <- gmfmc_landings_df %>%
  mutate(
    Series = if_else(
      COMMON_NAME %in% core_priority_species,
      "Core Priority Species Subset",
      "Other Managed GMFMC Finfish"
    )
  ) %>%
  group_by(LAND_YEAR, Series) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    .groups = "drop"
  )

# 2. Generate Comparative Stacked Area Plot
ggplot(comparison_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6, fill = Series)) +
  geom_area(alpha = 0.8, color = "black", linewidth = 0.3) +
  scale_fill_manual(values = c(
    "Core Priority Species Subset" = "#08519c", 
    "Other Managed GMFMC Finfish" = "#9ecae1"
  )) +
  scale_y_continuous(labels = comma_format(), limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Gulf of Mexico Managed Commercial Finfish Extraction (1972–Present)",
    subtitle = "Comparing Total GMFMC Managed Finfish Landings vs. Core Priority Species Subset",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    fill = "Species Selection Strategy",
    caption = "Data Source: NOAA SEFSC Commercial Landings Extraction MV. Excludes shellfish and non-GMFMC taxa."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey30", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```



### Unmanaged species

Let's go back to the clean_data and filter for the unmanaged species we want to plot. First some diagnostics.

```{r}
# Define exclusions (Managed GMFMC species + Shellfish + Broad Aggregates)
shellfish_and_broad <- c(
  "SHRIMP, NORTHERN BROWN", "SHRIMP, NORTHERN WHITE", "SHRIMP, NORTHERN PINK",
  "SHRIMP, ROYAL RED", "SHRIMP, ROCK", "LOBSTER, CARIBBEAN SPINY", "CRAB, GOLDEN DEEPSEA",
  "CRAB, BLUE", "CRAB, STONE", "OYSTER, EASTERN", "SNAIL, OTHER", "STARS, OTHER SEA"
)

# Filter for strictly unmanaged finfish taxa
unmanaged_diagnostics <- clean_data %>%
  filter(
    LAND_YEAR >= 1972,
    !COMMON_NAME %in% core_priority_species,
    !COMMON_NAME %in% shellfish_and_broad,
    !str_detect(COMMON_NAME, regex("UNCLASSIFIED|UNIDENTIFIED|OTHER|GENUS|FAMILY|NOS|SPEC", ignore_case = TRUE))
  ) %>%
  group_by(SPECIES_ITIS, COMMON_NAME) %>%
  summarise(
    TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE),
    TOTAL_NOMINAL_DOLLARS = sum(TOTAL_DOLLARS, na.rm = TRUE),
    YEARS_PRESENT = n_distinct(LAND_YEAR),
    .groups = "drop"
  ) %>%
  arrange(desc(TOTAL_LIVE_LBS))

# View top 30 unmanaged species by volume
print(head(unmanaged_diagnostics, 30))
```

Let's run a Pareto distribution & sensitivity analysis. Calculates the cumulative percentage of total unmanaged marine finfish volume accounted for by each species, ranks them from highest to lowest, and tests if the time series trend changes when using top 5, top 10, top 15, or top 25 species. Trying to remove species was like playing constant whack a mole. Random things just kept popping up. So a better approach was to make a whitelist of known major commercial marine finfish that are not managed. 

```{r}

# Explicit Whitelist: Major Unmanaged Commercial Marine Finfish in the Gulf of Mexico
unmanaged_marine_finfish_vector <- c(
  # Primary Coastal / Estuarine Finfish
  "MULLET, STRIPED", "MULLET, WHITE", "DRUM, BLACK", "CROAKER, ATLANTIC", 
  "SEATROUT, SPOTTED", "SEATROUT, SAND", "SHEEPSHEAD", "LADYFISH", 
  "POMPANO, FLORIDA", "SPOT", "KINGFISHES", "MOJARRAS", "BALLYHOOS", "PINFISH",
  
  # Coastal Pelagics & Baitfish (Non-Menhaden)
  "HERRING, ATLANTIC THREAD", "SARDINE, SPANISH", "JACK, CREVALLE", 
  "RUNNER, BLUE", "SCAD, ROUND", "SCADS", "BLUEFISH", "BUTTERFISH,GULF", 
  "SHAD, GIZZARD", "SHAD, AMERICAN", "HARVESTFISH", "LOOKDOWN"
)

# 2. Cumulative Landings Pareto Analysis
pareto_df <- clean_data %>%
  filter(
    LAND_YEAR >= 1972,
    COMMON_NAME %in% unmanaged_marine_finfish_vector
  ) %>%
  group_by(COMMON_NAME) %>%
  summarise(TOTAL_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TOTAL_LBS)) %>%
  mutate(
    RANK = row_number(),
    CUM_LBS = cumsum(TOTAL_LBS),
    CUM_PCT = (CUM_LBS / sum(TOTAL_LBS)) * 100
  )


# 3. Sensitivity Analysis: Plot trajectories using Top 5, Top 10, Top 15, vs Top 25
top5_spp  <- pareto_df %>% filter(RANK <= 5) %>% pull(COMMON_NAME)
top10_spp <- pareto_df %>% filter(RANK <= 10) %>% pull(COMMON_NAME)
top15_spp <- pareto_df %>% filter(RANK <= 15) %>% pull(COMMON_NAME)
top25_spp <- pareto_df %>% filter(RANK <= 25) %>% pull(COMMON_NAME)

sensitivity_df <- unmanaged_pool %>%
  filter(COMMON_NAME %in% top25_spp) %>%
  mutate(
    Cutoff = case_when(
      COMMON_NAME %in% top5_spp ~ "Top 5 Species",
      COMMON_NAME %in% top10_spp ~ "Top 10 Species",
      COMMON_NAME %in% top15_spp ~ "Top 15 Species",
      TRUE ~ "Top 25 Species"
    )
  ) %>%
  group_by(LAND_YEAR, Cutoff) %>%
  summarise(TOTAL_LIVE_LBS = sum(TOTAL_LIVE_POUNDS, na.rm = TRUE), .groups = "drop")

# Plot sensitivity curves
ggplot(sensitivity_df, aes(x = LAND_YEAR, y = TOTAL_LIVE_LBS / 1e6, color = Cutoff)) +
  geom_line(linewidth = 1.1) +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = seq(1972, 2025, by = 5)) +
  labs(
    title = "Sensitivity of Unmanaged Finfish Trajectory to Taxonomic Scope",
    subtitle = "Testing cumulative volume thresholds (Top 5 vs 10 vs 15 vs 25 species)",
    x = "Year",
    y = "Live Landings (Millions of Pounds)",
    color = "Inclusion Threshold"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```

Now we can get the unmanaged data

```{r}
# 1. Filter landings to the explicit unmanaged marine finfish list & adjust for inflation
dat_unmanaged <- clean_data %>%
  filter(
    LAND_YEAR >= 1972,
    COMMON_NAME %in% unmanaged_marine_finfish_vector
  ) %>%
  left_join(deflator_annual, by = "LAND_YEAR") %>%
  mutate(
    TOTAL_DOLLARS_2024 = TOTAL_DOLLARS * (target_deflator_2024 / annual_deflator),
    total_mil_lbs = TOTAL_LIVE_POUNDS / 1e6,
    total_mil_dollars_2024 = TOTAL_DOLLARS_2024 / 1e6
  )

# 2. Summarize annual totals
unmanaged_summary <- dat_unmanaged %>% 
  group_by(LAND_YEAR) %>%
  summarise(
    total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), 
    total_mil_dollars = sum(total_mil_dollars_2024, na.rm = TRUE), 
    .groups = "drop"
  ) %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)

# Save intermediate summary CSV
write_csv(dat_unmanaged, here("data/intermediate/unmanaged_species_landings_by_species.csv"))
write_csv(unmanaged_summary, here("data/intermediate/gulf_unmanaged_species_landings_1972_present.csv"))
```



Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function.
Replace "your_data" in the code below with whatever your dataframe is called.

```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Real value (2024 $ million)")
extent_names = c("Assessed finfish", "Assessed finfish")

formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(assessed_finfish, indicator_names, unit_names, extent_names)

#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Real value (2024 $ million)")
extent_names = c("Unassessed finfish", "Unassessed finfish")

formatted_unassessed_all = IEAnalyzeR::convert_cleaned_data(unmanaged_summary, indicator_names, unit_names, extent_names)

```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

```{r}
write.csv(formatted_assessed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unassessed_all, file = csv_filename2, row.names = F)
```

# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Managed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Top Unmanaged Finfish")
ggsave(filename = plot_filename2)
```



































































Ignore everything below this...





```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish = dat_assessed %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)



dat_unassessed_summary = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed_summary)

dat_unassessed_summary$total_mil_lbs = dat_unassessed_summary$total_live_lbs / 1e6
dat_unassessed_summary$total_mil_dollars = dat_unassessed_summary$total_dollars / 1e6

unassessed_finfish_all = dat_unassessed_summary %>% 
  group_by(LAND_YEAR) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
```




```{r}
# Now the unassessed species

# Keywords for reef fish families 
reef_family_keywords <- "PORGY|GRUNT|SNAPPER|GROUPER|SEA BASS|SEABASS|TRIGGERFISH|JACK|WRASSE|PARROTFISH|TILEFISH|HOGFISH|HIND|CONEY|GRAYSBY|TOMTATE|PINFISH"

# Filter and assign Guilds to Unassessed Reef Fish
unassessed_reef_fish <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>% 
  group_by(LAND_YEAR, COMMON_NAME, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_fish2 = unassessed_reef_fish %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef fish not confidential
write_csv(unassessed_reef_fish2, here("data/intermediate/gulf_unassessed_reef_fish_1972_present.csv"))

print(unique(unassessed_reef_fish$COMMON_NAME))


unassessed_reef_summary <- clean_data %>%
  filter(
    fmp_status == "Managed",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_family_keywords),
    # Must NOT be one of the 15 target species
    !tolower(COMMON_NAME) %in% target_species_exact
  ) %>%
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Jacks",
      TRUE ~ "Other Reef Fish"
    )
  ) %>%
  group_by(LAND_YEAR, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)

unassessed_reef_summary = unassessed_reef_summary %>% 
  filter(is_confidential != TRUE)

# Save unassessed reef summary
write_csv(unassessed_reef_summary, here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
```

```{r}
# OPTIONAL: Free up RAM by deleting the 40-million row raw dataset
rm(landings_raw, clean_data)
gc()
```





# 3. Save Formatted data as csv

This will save your data to the appropriate folder.

```{r}
write.csv(formatted_assessed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unassessed_all, file = csv_filename2, row.names = F)
```

# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.

```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.

```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Assessed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings & Value of Unassessed Finfish")
ggsave(filename = plot_filename2)
```

------------------------------------------------------------------------

Now that we have overall trends, we want to break down by species.
Starting with the assessed species.

```{r}
dat_assessed = read.csv(here("data/intermediate/gulf_15_target_assessed_species_1972_present.csv"))
head(dat_assessed)

dat_assessed$total_mil_lbs = dat_assessed$total_live_lbs / 1e6
dat_assessed$total_mil_dollars = dat_assessed$total_dollars / 1e6

assessed_finfish_species = dat_assessed %>% 
  group_by(LAND_YEAR, COMMON_NAME) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, COMMON_NAME, total_mil_lbs, total_mil_dollars)


for (i in unique(assessed_finfish_species$COMMON_NAME)) {
  
  species_data = assessed_finfish_species %>% 
    filter(COMMON_NAME == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_assessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_assessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```

Now let's do the same for the unassessed species.

```{r}
dat_unassessed = read.csv(here("data/intermediate/gulf_unassessed_reef_fish_guilds_1972_present.csv"))
head(dat_unassessed)

dat_unassessed$total_mil_lbs = dat_unassessed$total_live_lbs / 1e6
dat_unassessed$total_mil_dollars = dat_unassessed$total_dollars / 1e6

unassessed_finfish_species = dat_unassessed %>% 
  group_by(LAND_YEAR, guild) %>%
  summarize(total_mil_lbs = sum(total_mil_lbs, na.rm = TRUE), total_mil_dollars  = sum(total_mil_dollars, na.rm = TRUE), .groups = "drop") %>% 
  select(LAND_YEAR, guild, total_mil_lbs, total_mil_dollars)


for (i in unique(unassessed_finfish_species$guild)) {
  
  species_data = unassessed_finfish_species %>% 
    filter(guild == i) %>% 
    select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
  
  #Define header components for the data rows (ignore year). Fill in the blanks here.
  indicator_names = rep("Commercial landings", 2)
  unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
  extent_names = rep(i, 2)

  formatted_unassessed_finfish = IEAnalyzeR::convert_cleaned_data(species_data, indicator_names, unit_names, extent_names)
  
  # Sanitize species name for safe filename output
  clean_filename = gsub("[^A-Za-z0-9_]", "_", i)
  
  write.csv(formatted_unassessed_finfish, here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), row.names = F)
  
  data_obj<-IEAnalyzeR::data_prep(here(paste0("data/formatted/formatted_csvs/comm_land_value_", clean_filename, ".csv")), subind = "unit")
  
  saveRDS(data_obj, file = here(paste0("data/formatted/final_objects/comm_land_value_", clean_filename, ".rds")))
  
  IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE, sep_ylabs = TRUE, manual_title = paste0(clean_filename, "Commercial Landings & Value"))
ggsave(filename = here(paste0("figures/plots/comm_land_value_", clean_filename, ".png")))
  
}
```
