---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-07-21 by Carissa Gervasi. Required to have a login to Oracle to pull data, but non-confidential summaries are stored on Github. The documentation for analyzing commercial landings data can be found in the Gulf IEA google drive folder for the 2025 Gulf ESR. 

***Need to re-pull data and fix the unmanaged reef fish

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(ROracle)
library(keyring)
library(tidyverse)
library(scales)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "com_landings"

csv_filename1<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_managed_formatted.csv")) 
object_filename1<-here(paste0("data/formatted/final_objects/", root_name, "all_managed_object.rds"))
plot_filename1<-here(paste0("figures/plots/", root_name, "all_managed_plot.png"))

csv_filename2<-here(paste0("data/formatted/formatted_csvs/", root_name, "all_unmanaged_formatted.csv")) 
object_filename2<-here(paste0("data/formatted/final_objects/", root_name, "all_unmanaged_object.rds"))
plot_filename2<-here(paste0("figures/plots/", root_name, "all_unmanaged_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

```{r}
Sys.setenv(TZ = "UTC", ORA_SDTZ = "UTC")

# Define your username
my_username <- "CGERVASI"

# Connect using stored keyring password
con <- dbConnect(dbDriver("Oracle"), 
                 username = my_username,
                 password = key_get("SECPR_Oracle", my_username),
                 dbname = "SECPR")

# Run quick test query
#test_query <- "SELECT * FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV WHERE ROWNUM <= 1000"
#df <- dbGetQuery(con, test_query)

# SQL Query filtered at database level
query <- "SELECT LAND_YEAR, SPECIES_ITIS, COMMON_NAME, LIVE_POUNDS, DOLLARS, DEALER_ID, FLAG_FMP_SPP 
          FROM SEDAT.COM_LAND_EXTRACTION_MV@secdw_dblk.SFSC.NOAA.GOV
          WHERE LIVE_POUNDS > 0
          AND LAND_YEAR >= 1972
          AND REGION IN ('GOM', 'GOM INTL')"


# Fetch data
landings_raw <- dbGetQuery(con, query)


#Always close the connection when you're done
dbDisconnect(con)


```

# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).


SKIP THIS SECTION UNLESS YOU NEED TO PULL DATA FROM ORACLE AGAIN 

```{r}
# 1. Define Shellfish Filter & Exceptions
# ------------------------------------------------------------------------------
# The metadata specifies that this dataset is designed for finfish and explicitly warns against using it for food shrimp
to_remove <- "shrimp|crab|lobster|oyster|clam|scallop|mussel|squid|octopus|conch|crawfish|crayfish|urchin|snail|menhaden"
finfish_exceptions <- c("oyster toadfish", "shrimp scad")

# 2. Filter & Prepare Clean Working Data
# ------------------------------------------------------------------------------
clean_data <- landings_raw %>%
  filter(
    !is.na(LAND_YEAR),
    !str_detect(tolower(COMMON_NAME), to_remove) | 
      tolower(COMMON_NAME) %in% finfish_exceptions            
  ) %>%
  mutate(
    fmp_status = if_else(FLAG_FMP_SPP == 1, "Managed", "Unmanaged")
  )

# 3. Create Summary Dataset 1: Combined Totals (Managed vs. Unmanaged)
# ------------------------------------------------------------------------------
summary_combined <- clean_data %>%
  group_by(LAND_YEAR, fmp_status) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)                      # Rule of 3 check

summary_combined = summary_combined %>% 
  filter(is_confidential != TRUE)

# some checking 

# Check the top unmanaged finfish species by total volume
top_unmanaged_species <- clean_data %>%
  filter(fmp_status == "Unmanaged") %>%
  group_by(COMMON_NAME) %>%
  summarize(
    grand_total_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    grand_total_dollars = sum(DOLLARS, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(grand_total_lbs))

head(top_unmanaged_species, 10)
write.csv(top_unmanaged_species, here("data/intermediate/top_unmanaged_species.csv"), row.names = FALSE)


##### The unmanaged species are getting a little tricky. Perhaps we want to just look at unmanaged reef fish species...
# 1. Define Keywords for Reef Fish Families
reef_keywords <- "PORGY|GRUNT|SNAPPER|GROUPER|SEA BASS|SEABASS|TRIGGERFISH|JACK|WRASSE|PARROTFISH|TILEFISH|HOGFISH|HIND|CONEY|GRAYSBY|TOMTATE|PINFISH"

# 2. Re-assert your 15 managed species to exclude them explicitly
managed_15_regex <-   "cobia|grouper, gag| snapper, gray| triggerfish, gray| amberjack, greater| hogfish| mackerel, king| snapper, mutton| grouper, red| snapper, red|  scamp|mackerel, spanish| snapper, vermilion| grouper, yellowedge| snapper, yellowtail"

# 3. Filter and Guild Unmanaged Reef Fish
unmanaged_reef_fish <- clean_data %>%
  filter(
    fmp_status == "Unmanaged",
    # Must match a reef family keyword
    str_detect(toupper(COMMON_NAME), reef_keywords),
    # Must NOT be one of your 15 core managed species
    !str_detect(tolower(COMMON_NAME), managed_15_regex)
  ) %>%
  # Assign Ecological Guilds
  mutate(
    guild = case_when(
      str_detect(toupper(COMMON_NAME), "SNAPPER|GROUPER|HIND|CONEY|GRAYSBY") ~ "Unmanaged Snappers & Groupers",
      str_detect(toupper(COMMON_NAME), "PORGY|GRUNT|TOMTATE|PINFISH") ~ "Porgies & Grunts",
      str_detect(toupper(COMMON_NAME), "JACK|RUNNER") ~ "Unmanaged Jacks",
      TRUE ~ "Other Unmanaged Reef Fish"
    )
  )

# 4. Aggregate by Year and Guild
unmanaged_reef_summary <- unmanaged_reef_fish %>%
  group_by(LAND_YEAR, guild) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)



# 4. Create Summary Dataset 2: Individual Managed Species
# ------------------------------------------------------------------------------
summary_managed_spp <- clean_data %>%
  filter(fmp_status == "Managed") %>%
  group_by(LAND_YEAR, SPECIES_ITIS, COMMON_NAME) %>%
  summarize(
    total_live_lbs = sum(LIVE_POUNDS, na.rm = TRUE),
    total_dollars  = sum(DOLLARS, na.rm = TRUE),
    n_dealers      = n_distinct(DEALER_ID, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(is_confidential = n_dealers < 3)                      # Rule of 3 check

# Define your target list (handles spelling variations and shorthand names)
target_species_regex <- paste(c(
  "cobia", 
  "grouper, gag",                     
  "snapper, gray", 
  "triggerfish, gray", 
  "amberjack, greater", 
  "hogfish", 
  "mackerel, king", 
  "snapper, mutton", 
  "grouper, red", 
  "snapper, red", 
  "scamp",                    
  "mackerel, spanish", 
  "snapper, vermilion",       
  "grouper, yellowedge", 
  "snapper, yellowtail"
), collapse = "|")

# Filter your summarized managed species dataset
summary_target_managed <- summary_managed_spp %>%
  filter(
    str_detect(tolower(COMMON_NAME), target_species_regex),
    !str_detect(COMMON_NAME, "HOGFISH, SPANISH"),
    is_confidential == FALSE  
  )

# Verify which unique species names matched from your database
cat("Matched species in dataset:\n")
print(unique(summary_target_managed$COMMON_NAME))

# 5. Confidentiality Reporting & File Export
# ------------------------------------------------------------------------------
cat("Confidential rows in Combined Totals:", sum(summary_combined$is_confidential), "\n")
cat("Confidential rows in Managed Species breakdown:", sum(summary_managed_spp$is_confidential), "\n")

# Save lightweight summaries locally so you never have to pull 40M rows again!
write_csv(summary_combined, here("data/intermediate/gulf_combined_finfish_totals_1972_present.csv"))
write_csv(summary_managed_spp, here("data/intermediate/gulf_managed_species_landings_1972_present.csv"))

# OPTIONAL: Free up RAM by deleting the 40-million row raw dataset
rm(landings_raw, clean_data)
gc()
```

Start from here if you are just using the intermediate datasets. 
```{r}
dat_combined = read.csv(here("data/intermediate/gulf_combined_finfish_totals_1972_present.csv"))
head(dat_combined)

managed_finfish = dat_combined %>% 
  subset(fmp_status == "Managed")
managed_finfish$total_mil_lbs = managed_finfish$total_live_lbs / 1e6
managed_finfish$total_mil_dollars = managed_finfish$total_dollars / 1e6
managed_finfish = managed_finfish %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)

unmanaged_finfish = dat_combined %>% 
  subset(fmp_status == "Unmanaged") 
unmanaged_finfish$total_mil_lbs = unmanaged_finfish$total_live_lbs / 1e6
unmanaged_finfish$total_mil_dollars = unmanaged_finfish$total_dollars / 1e6
unmanaged_finfish = unmanaged_finfish %>% 
  select(LAND_YEAR, total_mil_lbs, total_mil_dollars)
```


Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.

```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
extent_names = c("Managed finfish", "Managed finfish")

formatted_managed_finfish = IEAnalyzeR::convert_cleaned_data(managed_finfish, indicator_names, unit_names, extent_names)


#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = rep("Commercial landings", 2)
unit_names = c("Whole weight (million lbs)", "Million dollars ($)")
extent_names = rep("Unmanaged finfish", 2)

formatted_unmanaged_finfish = IEAnalyzeR::convert_cleaned_data(unmanaged_finfish, indicator_names, unit_names, extent_names)
```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.
```{r}
write.csv(formatted_managed_finfish, file = csv_filename1, row.names = F)
write.csv(formatted_unmanaged_finfish, file = csv_filename2, row.names = F)
```


# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.
```{r}
data_obj1<-IEAnalyzeR::data_prep(csv_filename1, subind = "unit")
data_obj2<-IEAnalyzeR::data_prep(csv_filename2, subind = "unit")
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.
```{r}
saveRDS(data_obj1, file = object_filename1)
saveRDS(data_obj2, file = object_filename2)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj1, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings and Value of Managed Finfish")
ggsave(filename = plot_filename1)

IEAnalyzeR::plot_fn_obj(df_obj = data_obj2, trend = TRUE, sep_ylabs = TRUE, manual_title = "Commercial Landings and Value of Unmanaged Finfish")
ggsave(filename = plot_filename2)
```

