---
title: "Metric Analysis Code"
format: html
execute: 
  echo: true
  warning: false
editor: 
  markdown: 
    wrap: sentence
---

File created on 2026-09-30

# 0. Setup

Add any packages that are needed for analysis in this code chunk.

```{r packages}
library(IEAnalyzeR)
library(here)
library(ggplot2)
library(ROracle)
library(keyring)
library(tidyverse)
library(scales)
library(stringr)
```

File Naming Setup.
!! Auto generated-Do Not Change !!

```{r}

root_name<- "rec_data"

csv_filename1<-here(paste0("data/formatted/formatted_csvs/", root_name, "_managed_spp_formatted.csv")) 
object_filename1<-here(paste0("data/formatted/final_objects/", root_name, "_managed_spp_object.rds"))
plot_filename1<-here(paste0("figures/plots/", root_name, "_managed_spp_plot.png"))

csv_filename2<-here(paste0("data/formatted/formatted_csvs/", root_name, "_unmanaged_spp_formatted.csv")) 
object_filename2<-here(paste0("data/formatted/final_objects/", root_name, "_unmanaged_spp_object.rds"))
plot_filename2<-here(paste0("figures/plots/", root_name, "_unmanaged_spp_plot.png"))

```

# 1. Read Data

Pull data from its source: Manual data: data/unformatted data. Automated data: Add script for data call (API, package, etc.). Confidential data: Store locally in the confidential data folder - This folder is excluded using gitignore and will not push to the GitHub repo. If intermediate data (shapefiles etc.) are needed, please put them in data\>intermediate - Filename should use the syntax rootname_descriptivename

Guidance for the rec data: https://sefsc.github.io/SEFSC-dolphin-analyses/MRIP.html

```{r}
Sys.setenv(TZ = "UTC", ORA_SDTZ = "UTC")

# Define your username
my_username <- "CGERVASI"

# Connect using stored keyring password
con <- dbConnect(dbDriver("Oracle"), 
                 username = my_username,
                 password = key_get("SECPR_Oracle", my_username),
                 dbname = "SECPR")
```

```{r}
# Quick test loading the data. Remember to be on the VPN.

ACL.test = dbGetQuery(con,
                      paste0("select *
                        from rdi.V_REC_ACL_MRIP_FES_IMP@secapxdv_dblk.sfsc.noaa.gov WHERE ROWNUM <= 1000" ))
```

```{r}
#Now run the real query to get the entire table (takes a few min)

ACL.raw <- dbGetQuery(con, "SELECT *
                     FROM rdi.V_REC_ACL_MRIP_FES_IMP@secapxdv_dblk.sfsc.noaa.gov")
```

```{r}
# close the connection

dbDisconnect(con)
```


# 2. Clean data and create time series csv

Transform the data to fit the IEA data format.
For more info on IEA data format go to the IEAnalyzeR vignette [HERE](https://gulf-iea.github.io/IEAnalyzeR/articles/How_to_use_IEAnalyzeR.html).
Once data are formatted with time (annual or monthly) as column 1 and metric values in the remaining columns, you can use the function convert_cleaned_data to convert your csv into a format that can be read by the data_prep function. Replace "your_data" in the code below with whatever your dataframe is called.

```{r}
head(ACL.raw)
str(ACL.raw)

#first we will subset to the Gulf. That is SUB_REG = 7. 

ACL.gulf = ACL.raw %>% 
  subset(SUB_REG == 7)

#Now let's just get a list of all the species in the dataset and their total landings
gulf_specs = ACL.gulf %>% 
  group_by(NEW_COM) %>% 
  summarize(
    GOM_LABEL = first(GOM_LABEL),
    total_landings = sum(LBSEST_SECWWT, na.rm = TRUE)
  ) %>% 
  arrange(GOM_LABEL, NEW_COM)

#save this to look at more closely
write_csv(gulf_specs, here("data/intermediate/gulf_rec_specs_list.csv"))
```


```{r}
#Define header components for the data rows (ignore year). Fill in the blanks here.
indicator_names = c("")
unit_names = c("")
extent_names = c("")

formatted_data = IEAnalyzeR::convert_cleaned_data(your_data, indicator_names, unit_names, extent_names)
```

# 3. Save Formatted data as csv

This will save your data to the appropriate folder.
```{r}
write.csv(formatted_data, file = csv_filename, row.names = F)
```


# 4. Create Data_Prep object

Please use your formatted csv to create a "data_prep" object.
For more info on the data_prep function see the vignette linked above.
```{r}
data_obj<-IEAnalyzeR::data_prep(csv_filename)
```

# 5. Save Formatted data_prep object

This will save your data to the appropriate folder.
```{r}
saveRDS(data_obj, file = object_filename)
```

# 6. Preview Plot

Use the IEAnalyzeR plotting function to preview the data.
This will not necessarily be the final figure used in reports.
For more info on the plot_fn_obj function go HERE

```{r}
IEAnalyzeR::plot_fn_obj(df_obj = data_obj, trend = TRUE)
```

# 7. Save plot

This will save the plot to the correct folder.
Adjust height & width using (height=, width=, unit="in") if needed.

```{r}
ggsave(filename = plot_filename)
```
